# Indexing data

## Indexing introduction

What is indexing? In programming lingo it is the process of getting particular
values out of the data. For example, we want to get the 5th row of our data frame.
Or all the rows between 5 and 20. Or the third column. This may not be particularly
useful by itself, but we can index using the results from expressions.

Indexing in R is very useful and necessary, but confusing. I'll stick to the very
basics. If you need to go further, check ?\`[\`.

## Indexing rows

How do I get the 5th row?

```{r}
# let's get barley back
# we can automatically convert strings to factors if we ask R to...
# which is what we did by hand in `reading_and_viewing.Rmd`
barley <- read.csv(file = "data/barley.csv", stringsAsFactors = TRUE)

# get the 5th row
barley[5,]

```
Even though that's a short amount of code, there is a bit to break down. We index
into data frames using `[`. Generally it works like this: `x[i, j]`, where `i` is
the row index and `j` is the column index. These can be numbers, or characters if
the rows or columns are named. The comma is necessary after the 5 in our example.
Without it, we'd select the 5th column!

Say we want the 5th row and 2nd column.

```{r}
barley[5, 2]
```

Will return the value. If we want to keep the data frame structure, you have to 
specify that you don't want to drop the two dimensional structure.

```{r}
barley[5, 2, drop = FALSE]
```
See the difference? I don't know why R did not enable this by default but hey ho.

One more important and common indexing rule is `$`. `$` gets the column by name, 
not number.

```{r}
# extract yield, returning a vector
yield <- barley$yield

# this is equivalent to 
yield <- barley[["yield"]]
```

The double `[[` extracts a single element from a list/data frame. These both return
vectors of `yield`, so we can index those further.

```{r}
# should be equivalent to the above where we indexed into barley.
yield[5]
```

I told you it was a bit confusing.

## Useful indexing

It's all very well indexing into certain things manually, but the true power of 
indexing happens when we index based on the result from an expression. Using our
new found indexing powers, we can start filtering usefully.

Say we want to return all the rows where yield is greater than 50 bushels per acre.
We want to create a logical vector of all the places where this is true first.

```{r}
yield_50 <- barley$yield > 50
yield_50
```
So again, a bit to break down here. We first get the yield vector (`barley$yield`).
We then use the greater than operator to ask for every element in the yield vector
'is this element greater than 50?'. So `>` is a vectorised function - as we touched on
in a previous tutorial.

Now we can use this vector to subset our data frame!

```{r}
barley[yield_50, ]
```

We pass the logical vector into our square braces and remember to add a comma after
as we are indexing into the rows.

We might want to only extract rows of a certain variety. No problem, we know how 
to do that now.

```{r}
# the variety I want
variety_we_want <- "Trebi"

# the column and logical test
trebi <- barley$variety == variety_we_want

# and index our data frame
trebi_data <- barley[trebi, ]

# return the data
trebi_data
```

So there are the 12 rows which relate to yield for the `Trebi` variety. If we wanted
the yields associated with `Trebi`, that's easy too.

```{r}
# using our prior knowledge
trebi_data_yield_dollar <- trebi_data$yield

# or
trebi_data_yield_braces <- trebi_data[["yield"]]

# test they are equal
all.equal(trebi_data_yield_dollar, trebi_data_yield_braces)
```




