---
title: "Indexing Data"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 60)

barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)
```

## What is indexing?

Indexing is the process of extracting particular values from data. For example:
- Get the 5th row
- All rows between 5 and 20
- The third column
- All rows where yield > 50

Indexing in R is very useful but can be confusing. We will stick to the very
basics. If you need to go further, check out `` ?`[` ``.

---

## Indexing rows and columns

The general syntax is `x[i, j]` where `i` is the row index and `j` is the
column index.

```{r demo-read, echo=TRUE}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)

# get the 5th row (note the comma! it means "all columns")
barley[5, ]
```

```{r demo-row-col, echo=TRUE}
# 5th row, 2nd column - returns a single value
barley[5, 2]

# keep the data frame structure with drop = FALSE
barley[5, 2, drop = FALSE]
```

The `$` operator gets a column by name and returns it as a vector:

```{r demo-dollar, echo=TRUE}
# extract yield as a vector
yield <- barley$yield

# equivalent using double brackets
yield2 <- barley[["yield"]]

# they are identical
all.equal(yield, yield2)
```

And you can index further into the resulting vector:

```{r demo-vector-index, echo=TRUE}
# get the 5th yield value
yield[5]
```

### Your turn: basic indexing

```{r ex-index-basic, exercise=TRUE, exercise.lines=8}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)

# 1. Get the 52nd row of barley
barley[___, ]

# 2. Get the 3rd column by number
barley[, ___]
```

```{r ex-index-basic-hint}
# For rows: barley[row_number, ]
# For columns: barley[, column_number]
```

```{r ex-index-basic-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)
barley[52, ]
barley[, 3]
```

```{r ex-index-name, exercise=TRUE, exercise.lines=6}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)

# Get the "variety" column by name (two ways: $ and [[]])
barley$___
barley[[___]]
```

```{r ex-index-name-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)
barley$variety
barley[["variety"]]
```

```{r quiz-indexing-1, echo=FALSE}
quiz(
  question("What does `barley[5, ]` return?",
    answer("The 5th row, all columns", correct = TRUE),
    answer("The 5th column, all rows"),
    answer("The first 5 rows"),
    answer("The value in row 5, column 5"),
    allow_retry = TRUE
  ),
  question("What does `barley[, 3]` return?",
    answer("The 3rd column, all rows", correct = TRUE),
    answer("The 3rd row, all columns"),
    answer("The first 3 columns"),
    answer("An error"),
    allow_retry = TRUE
  )
)
```

---

## Filtering with logical indexing

The real power of indexing comes from filtering with **logical expressions**.

```{r demo-filter, echo=TRUE}
# create a logical vector: TRUE where yield > 50
yield_50 <- barley$yield > 50
head(yield_50)  # a vector of TRUE/FALSE

# use it to subset the data frame
barley_high_yield <- barley[yield_50, ]
nrow(barley_high_yield)
```

We can also filter by a specific value:

```{r demo-filter-variety, echo=TRUE}
# rows where variety is "Trebi"
trebi_rows <- barley$variety == "Trebi"
trebi_data <- barley[trebi_rows, ]
trebi_data
```

Combine multiple conditions with `&` (AND) and `|` (OR):

```{r demo-filter-combined, echo=TRUE}
# Trebi rows AND yield over 50
trebi_and_high <- trebi_rows & yield_50
barley[trebi_and_high, ]
```

### Your turn: filtering

```{r ex-filter-1, exercise=TRUE, exercise.lines=8}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)

# Filter to rows where yield is less than 25 bushels per acre
low_yield <- barley$yield ___ 25
barley[low_yield, ]
```

```{r ex-filter-1-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)
low_yield <- barley$yield < 25
barley[low_yield, ]
```

```{r ex-filter-2, exercise=TRUE, exercise.lines=10}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)

# Filter to rows from year 1932 AND site "Duluth"
year_1932 <- barley$year ___ 1932
duluth    <- barley$site ___ "Duluth"

barley[year_1932 ___ duluth, ]
```

```{r ex-filter-2-hint}
# == for equality, & for AND
```

```{r ex-filter-2-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)
year_1932 <- barley$year == 1932
duluth    <- barley$site == "Duluth"
barley[year_1932 & duluth, ]
```

```{r quiz-filtering, echo=FALSE}
quiz(
  question("What operator checks equality in R?",
    answer("`==`", correct = TRUE),
    answer("`=`"),
    answer("`!=`"),
    answer("`<-`"),
    allow_retry = TRUE
  ),
  question("What does `&` do when combining two logical vectors?",
    answer("Returns TRUE only where both are TRUE (element-wise AND)", correct = TRUE),
    answer("Returns TRUE where either is TRUE"),
    answer("Adds the two vectors together"),
    answer("Combines them into a list"),
    allow_retry = TRUE
  )
)
```

---

## Sorting data

Sorting requires generating a vector of indices that describe the desired order.
The `order()` function does this.

```{r demo-sort, echo=TRUE}
# generate sorted indices (ascending by yield)
sorted_idx <- order(barley$yield)

# use those indices to reorder the data frame
sorted_barley <- barley[sorted_idx, ]
head(sorted_barley)
```

For descending order, wrap with `rev()`:

```{r demo-sort-desc, echo=TRUE}
sorted_barley_desc <- barley[rev(sorted_idx), ]
head(sorted_barley_desc)
```

Find the row with the highest or lowest value:

```{r demo-which, echo=TRUE}
# which.max returns the INDEX of the maximum value
max_idx <- which.max(barley$yield)
barley[max_idx, ]
```

### Your turn: sorting

```{r ex-sort, exercise=TRUE, exercise.lines=8}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)

# Sort the data frame by year (ascending)
year_order <- order(___)
barley[year_order, ]
```

```{r ex-sort-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)
year_order <- order(barley$year)
barley[year_order, ]
```

```{r ex-which-min, exercise=TRUE, exercise.lines=6}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)

# Get the row corresponding to the MINIMUM yield
min_idx <- ___(barley$yield)
barley[min_idx, ]
```

```{r ex-which-min-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)
min_idx <- which.min(barley$yield)
barley[min_idx, ]
```

## Final quiz

```{r quiz-indexing-final, echo=FALSE}
quiz(
  question("How do you sort a data frame by a column in descending order?",
    answer("`df[rev(order(df$col)), ]`", correct = TRUE),
    answer("`df[order(df$col, decreasing = TRUE), ]`", correct = TRUE,
           message = "Also correct! There is more than one way."),
    answer("`sort(df, by = df$col)`"),
    answer("`df[sort(df$col), ]`"),
    allow_retry = TRUE
  ),
  question("What does `which.max(x)` return?",
    answer("The INDEX of the maximum value in x", correct = TRUE),
    answer("The maximum value in x"),
    answer("All values greater than the max"),
    answer("A logical vector"),
    allow_retry = TRUE
  )
)
```
