---
title: "Reading and Viewing Data"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 60)

# load barley for exercises
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)
```

## Reading in data

We will be working with yield data from a Minnesota barley trial. Let's see how
we read data into R, and then get a sense of what the data looks like.

Reading in data almost always requires a **file path** that points to where the
file lives on your computer — or a URL pointing to it online. We read CSV files
with `read.csv()`.

```{r demo-read, echo=TRUE}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"

barley <- read.csv(barley_url)
```

That's it! R stores the data internally as a `data.frame`.

> **Recommendation:** Always work with plain text files (CSV, TSV) rather than
> Excel files. Excel's formatting gets in the way of reproducibility — if you
> need to encode extra information, add it as a column.

## Viewing data

We should always look at the data structure before going further.

```{r demo-str, echo=TRUE}
# str() is your best friend - always call it on new data
str(barley)
```

We can see there are 5 columns: types `int`, `num`, and `chr`. In most cases
we want character columns to be **factors**.

```{r demo-factor, echo=TRUE}
# convert variety and site columns to factors
barley$variety <- as.factor(barley$variety)
barley$site    <- as.factor(barley$site)

str(barley)
```

Now we can see how many unique levels each factor has.

```{r demo-head, echo=TRUE}
# head() shows the first 6 rows by default
head(barley)
```

### Your turn: explore the structure

```{r ex-str, exercise=TRUE, exercise.lines=8}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)

# 1. How many rows and columns does barley have?
#    Hint: nrow() and ncol(), or dim()
nrow(___)
ncol(___)
```

```{r ex-str-hint}
# nrow(barley) gives number of rows
# ncol(barley) gives number of columns
# dim(barley) gives both at once
```

```{r ex-str-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)
nrow(barley)
ncol(barley)
dim(barley)
```

```{r ex-tail, exercise=TRUE, exercise.lines=6}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)

# How would you see the LAST few rows? Hint: the opposite of head()
___(barley)
```

```{r ex-tail-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)
tail(barley)
```

```{r quiz-reading, echo=FALSE}
quiz(
  question("What function reads a CSV file into R?",
    answer("`read.csv()`", correct = TRUE),
    answer("`load.csv()`"),
    answer("`import()`"),
    answer("`open.file()`"),
    allow_retry = TRUE
  ),
  question("What does `str()` stand for?",
    answer("Structure", correct = TRUE),
    answer("String"),
    answer("Store"),
    answer("Summary table"),
    allow_retry = TRUE
  ),
  question("Why convert character columns to factors?",
    answer("So R knows the set of valid values; helps with statistics and plotting", correct = TRUE),
    answer("To save memory"),
    answer("Because str() requires it"),
    answer("Characters can't be used in data frames"),
    allow_retry = TRUE
  )
)
```

---

## Column names and row names

A data frame always has column and row names.

```{r demo-colnames, echo=TRUE}
cat("Original names:", paste(colnames(barley), collapse = ", "))
```

You can change them by assigning a vector of the same length:

```{r demo-rename, echo=TRUE}
colnames(barley) <- c("Row Names", "Yield", "Variety", "Year", "Site")
cat("Altered names:", paste(colnames(barley), collapse = ", "))

# restore original names for subsequent exercises
colnames(barley) <- c("X", "yield", "variety", "year", "site")
```

### Your turn: column names

```{r ex-colnames, exercise=TRUE, exercise.lines=10}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)

# 1. Print the column names of barley
colnames(___)

# 2. Rename the columns to: "ID", "Yield_bushels", "Variety", "Year", "Site"
colnames(barley) <- c(___)
head(barley)
```

```{r ex-colnames-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)
colnames(barley)
colnames(barley) <- c("ID", "Yield_bushels", "Variety", "Year", "Site")
head(barley)
```

## Reading TSV and other formats

`read.csv()` is just a shortcut for `read.table()` with `sep = ","`. You can
read tab-separated files (TSV) like this:

```{r demo-tsv, echo=TRUE, eval=FALSE}
# Reading a TSV file
my_data <- read.table("file.tsv", sep = "\t", header = TRUE)

# Or equivalently
my_data <- read.csv("file.tsv", sep = "\t")
```

```{r quiz-formats, echo=FALSE}
quiz(
  question("How would you read a tab-separated values (TSV) file?",
    answer('`read.csv("file.tsv", sep = "\\t")`', correct = TRUE),
    answer('`read.tsv("file.tsv")`'),
    answer('`load.tsv("file.tsv")`'),
    answer('`read.table("file.tsv")`  (without any extra arguments)',
           message = "Close! But you also need `sep = '\\t'` and `header = TRUE`."),
    allow_retry = TRUE
  )
)
```
