# Writing data

So far we have either created objects in R or loaded them into R using `read.csv`.
It's essential to also write out files from R onto our computer. This allows us
to open the file in other applications.

## Writing CSV files

First we can start by loading in our data again.

```{r}
barley <- read.csv("https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv", stringsAsFactors = TRUE)
# barley <- read.csv("data/barley.csv", stringsAsFactors = TRUE)
```

And we can filter this data in some interesting way. I am interested this time in
all yields less than the mean yield across the whole data set.

```{r}
# compute the mean
# note that `mean` is actually the name of a function
# so don't overwrite it!!
global_mean <- mean(barley$yield)

# and filter our data, just like we did in `indexing`
below_global_mean <- barley[barley$yield < global_mean, ]

# and let's look at the structure instead of printing everything out.
str(below_global_mean)
```
Great, we've filtered the data and have only got 68 observations or rows. We can 
write this to a file. The `write.csv` function is our friend if we want to write
a CSV specifically. It minimally takes `x` which is the R object to write and 
`file` which is the location of where we want to write the file.

```{r}
# simply call the write.csv function.
write.csv(x = below_global_mean, file = "./data/below_global_mean.csv")
```

If we check the `data` folder, it should now have another CSV file written in to it.

## Over to you

1. Read in `barley` and write out the first 10 lines to `./data/barley_first_10_lines.csv`
2. Can you convert `barley.csv` to `barley.tsv` changing the commas to tabs?





