---
title: "Writing Data"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 60)

barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)
```

## Writing data out of R

So far we have either created objects in R or loaded them using `read.csv()`.
It's essential to also **write files** back to disk so you can open them in
other applications or share them with colleagues.

## Writing CSV files

First, load and filter the data in some meaningful way. Here we keep only the
rows where yield is below the overall mean.

```{r demo-filter, echo=TRUE}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)

# compute overall mean yield
global_mean <- mean(barley$yield)

# filter to below-average rows
below_mean <- barley[barley$yield < global_mean, ]

str(below_mean)
```

Now write this filtered dataset out:

```{r demo-write, echo=TRUE, eval=FALSE}
# write.csv saves a data frame to a CSV file
# x    = the R object to write
# file = where to save it
write.csv(x = below_mean, file = "data/below_global_mean.csv")
```

> In this interactive tutorial the file system is sandboxed, so we write to a
> temporary file. On your own machine you would use a real path like `"data/below_global_mean.csv"`.

```{r demo-write-temp, echo=TRUE}
# write to a temp file and read it back to verify
tmp <- tempfile(fileext = ".csv")
write.csv(x = below_mean, file = tmp)

# read it back
verify <- read.csv(tmp)
cat("Rows written:", nrow(verify), "\n")
head(verify)
```

## Controlling the output

Two common options to know:

```{r demo-options, echo=TRUE, eval=FALSE}
# row.names = FALSE avoids writing a redundant row number column
write.csv(x = below_mean, file = "output.csv", row.names = FALSE)

# write.table gives full control over separators
write.table(x = below_mean, file = "output.tsv",
            sep = "\t", row.names = FALSE)
```

### Your turn: write the first 10 rows

```{r ex-write-10, exercise=TRUE, exercise.lines=12}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)

# 1. Extract the first 10 rows using head()
first_10 <- head(___, n = ___)

# 2. Write them to a temp file (no row names)
tmp <- tempfile(fileext = ".csv")
write.csv(x = ___, file = tmp, row.names = ___)

# 3. Read it back and print it to verify
read.csv(tmp)
```

```{r ex-write-10-hint}
# head(barley, n = 10) gives first 10 rows
# row.names = FALSE avoids an extra column
```

```{r ex-write-10-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)
first_10 <- head(barley, n = 10)
tmp <- tempfile(fileext = ".csv")
write.csv(x = first_10, file = tmp, row.names = FALSE)
read.csv(tmp)
```

### Your turn: convert CSV to TSV

```{r ex-write-tsv, exercise=TRUE, exercise.lines=12}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)

# Write barley as a TSV (tab-separated) file
# Hint: use write.table() with sep = "\t"
tmp_tsv <- tempfile(fileext = ".tsv")
write.table(x = barley, file = tmp_tsv, sep = ___, row.names = FALSE)

# Read it back using read.table (or read.csv with sep = "\t")
back <- read.table(tmp_tsv, sep = ___, header = ___)
head(back)
```

```{r ex-write-tsv-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)
tmp_tsv <- tempfile(fileext = ".tsv")
write.table(x = barley, file = tmp_tsv, sep = "\t", row.names = FALSE)
back <- read.table(tmp_tsv, sep = "\t", header = TRUE)
head(back)
```

## Quiz

```{r quiz-writing, echo=FALSE}
quiz(
  question("Which function writes a data frame to a CSV file?",
    answer("`write.csv()`", correct = TRUE),
    answer("`save.csv()`"),
    answer("`export()`"),
    answer("`write.table()` only"),
    allow_retry = TRUE
  ),
  question("What does `row.names = FALSE` do in `write.csv()`?",
    answer("Suppresses the automatic row number column in the output file",
           correct = TRUE),
    answer("Removes the column names"),
    answer("Makes the file smaller by removing metadata"),
    answer("Nothing — it is the default"),
    allow_retry = TRUE
  ),
  question("How would you write a TSV (tab-separated) file?",
    answer('`write.table(df, "out.tsv", sep = "\\t")`', correct = TRUE),
    answer('`write.csv(df, "out.tsv")`'),
    answer('`write.tsv(df, "out.tsv")`'),
    answer('`export.tsv(df, "out.tsv")`'),
    allow_retry = TRUE
  )
)
```
