# Functions

We have already encountered many functions in this book. We will break
down functions further in this short note.

## Ordinary functions

We will work on a simple data type, the vector. Functions are called by
wrapping an object in the function call. For example to take the mean of
a vector of numbers:

```{r}
# make some dummy data
data <- c(12, 30, 2, 34, 9, 8, 64, 12, 3, 99)

# take the mean
# we call the function by naming the function and opening parentheses
# around the object
mean(data)
```

Functions can have parameters. These are extra arguments which are added
to the function call that modify the behavior of the function. In `mean`
we see there is another argument called `na.rm`. Let's investigate.

```{r}
# append an NA - or Not Available. It's missing data.
# we can tack a bit of missing data on the end using
# the c() function.
na_data <- c(data, NA)

# now call mean
mean(na_data)
```

Hang on - we got an `NA` back? This is because the default
implementation of `mean` will always return `NA` if any `NA`'s are
present in the data! We have to tell R to remove them explicitly.

```{r}
# pass TRUE to the na.rm parameter to remove those pesky NA values
mean_na_rm <- mean(na_data, na.rm = TRUE)
```

We can also call multiple functions at the same time. Say we have two
data sets and we want to get the mean of each of them and then sum the
results, we can do that easily.

```{r}
# data as above
# data 2 below
data_2 <- c(102, 3, 2, 4, 9, 2, 1, 12, 97)

# sum and mean each data individually
sum(mean(data), mean(data_2))
```

## Special functions

R has many functions you may not think are functions. One is `+`.

```{r}
# okay that's fine
two <- 1 + 1

also_two <- `+`(1, 1)

all.equal(two, also_two)
```

We can do this wizardry as `+` is a function! Usually it's called with
one argument either side of the function. But we can 'capture' the
function by putting it in backticks, and pass the arguments like we
would call any other function.

Another weird function is `[`.

```{r}
# okay, fine!
first <- data[1]

# huh??
also_first <- `[`(data, 1)

all.equal(first, also_first)
```

The square braces in all their forms are functions, and can be called
like normal functions. You wouldn't want to do this though because it
wouldn't be good programming style.
