---
title: "Using Functions on Data Frames"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 60)

cats_url   <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/cats.csv"
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"

cats   <- read.csv(cats_url,   stringsAsFactors = TRUE)
barley <- read.csv(barley_url, stringsAsFactors = TRUE)
```

## Overview

Given that we have already looked at indexing in data frames, we have already
used functions on data frames! Here we will look at higher-level functions that
let us apply operations across columns, rows, or groups without writing loops.

The four core functions are:

| Function   | What it does |
|-----------|------|
| `lapply()` | Apply a function over a list/columns → returns a **list** |
| `sapply()` | Same as lapply, but **simplifies** the output |
| `apply()`  | Apply over rows or columns of a **matrix/data frame** |
| `tapply()` | Apply over **groups** defined by a factor |

---

## Functions on multiple columns: `lapply` and `sapply`

```{r demo-load-cats, echo=TRUE}
cats_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/cats.csv"
cats <- read.csv(cats_url, stringsAsFactors = TRUE)
str(cats)
```

Compute the mean of multiple columns at once using `lapply`:

```{r demo-lapply, echo=TRUE}
# select the two numeric columns
numeric_cols <- cats[, c("Bwt", "Hwt")]

# apply mean to each column — returns a list
lapply(X = numeric_cols, FUN = mean)
```

`sapply` simplifies the output to a named vector or matrix:

```{r demo-sapply, echo=TRUE}
sapply(X = numeric_cols, FUN = mean)
```

Pass a custom function to compute multiple statistics at once:

```{r demo-sapply-custom, echo=TRUE}
sapply(
  X   = numeric_cols,
  FUN = function(col) c(mean = mean(col), sd = sd(col), n = length(col))
)
```

### Your turn: lapply and sapply

```{r ex-lapply, exercise=TRUE, exercise.lines=12}
cats_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/cats.csv"
cats <- read.csv(cats_url, stringsAsFactors = TRUE)

numeric_cols <- cats[, c("Bwt", "Hwt")]

# 1. Use sapply to get the MEDIAN of each numeric column
sapply(X = numeric_cols, FUN = ___)

# 2. Use sapply with a custom function to get both min and max
sapply(X = numeric_cols, FUN = function(col) c(min = ___, max = ___))
```

```{r ex-lapply-solution}
cats_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/cats.csv"
cats <- read.csv(cats_url, stringsAsFactors = TRUE)
numeric_cols <- cats[, c("Bwt", "Hwt")]
sapply(X = numeric_cols, FUN = median)
sapply(X = numeric_cols, FUN = function(col) c(min = min(col), max = max(col)))
```

```{r quiz-apply-1, echo=FALSE}
quiz(
  question("What is the difference between `lapply` and `sapply`?",
    answer("`lapply` always returns a list; `sapply` tries to simplify to a vector or matrix",
           correct = TRUE),
    answer("`sapply` is for lists only; `lapply` works on any object"),
    answer("They are identical — one is just an alias"),
    answer("`lapply` is faster than `sapply`"),
    allow_retry = TRUE
  )
)
```

---

## Functions over rows: `apply`

In R, operations are naturally column-oriented. But `apply()` with `MARGIN = 1`
operates over rows:

```{r demo-apply-rows, echo=TRUE}
cats_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/cats.csv"
cats <- read.csv(cats_url, stringsAsFactors = TRUE)

# convert body weight to grams to match heart weight units
bwt_hwt <- data.frame(
  Bwt_g = cats$Bwt * 1000,
  Hwt_g = cats$Hwt
)

# MARGIN = 1 means apply over rows
# here: sum Bwt and Hwt for each animal
row_sums <- apply(X = bwt_hwt, MARGIN = 1, FUN = sum)
head(row_sums)
```

`MARGIN = 2` applies over columns (equivalent to `sapply`):

```{r demo-apply-cols, echo=TRUE}
apply(X = bwt_hwt, MARGIN = 2, FUN = mean)
```

### Your turn: apply

```{r ex-apply, exercise=TRUE, exercise.lines=14}
cats_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/cats.csv"
cats <- read.csv(cats_url, stringsAsFactors = TRUE)

bwt_hwt <- data.frame(Bwt_g = cats$Bwt * 1000, Hwt_g = cats$Hwt)

# 1. Compute the range (max - min) of each ROW using apply
#    Hint: diff(range(x)) gives the range of a vector
row_ranges <- apply(X = bwt_hwt, MARGIN = ___, FUN = function(x) diff(range(x)))
head(row_ranges)

# 2. Compute the standard deviation of each COLUMN
apply(X = bwt_hwt, MARGIN = ___, FUN = sd)
```

```{r ex-apply-solution}
cats_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/cats.csv"
cats <- read.csv(cats_url, stringsAsFactors = TRUE)
bwt_hwt <- data.frame(Bwt_g = cats$Bwt * 1000, Hwt_g = cats$Hwt)
row_ranges <- apply(X = bwt_hwt, MARGIN = 1, FUN = function(x) diff(range(x)))
head(row_ranges)
apply(X = bwt_hwt, MARGIN = 2, FUN = sd)
```

---

## Functions over groups: `tapply`

`tapply()` applies a function to subsets of a vector defined by a grouping
factor. This is very common in biological analyses.

```{r demo-tapply, echo=TRUE}
cats_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/cats.csv"
cats <- read.csv(cats_url, stringsAsFactors = TRUE)

# mean body weight by sex
tapply(X = cats$Bwt, INDEX = cats$Sex, FUN = mean)
```

```{r demo-tapply-barley, echo=TRUE}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)

# mean yield per site
tapply(X = barley$yield, INDEX = barley$site, FUN = mean)
```

### Your turn: tapply

```{r ex-tapply, exercise=TRUE, exercise.lines=12}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)

# 1. Compute the MAXIMUM yield per variety
tapply(X = ___, INDEX = barley$variety, FUN = ___)
```

```{r ex-tapply-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)
tapply(X = barley$yield, INDEX = barley$variety, FUN = max)
```

```{r ex-tapply-2, exercise=TRUE, exercise.lines=12}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)

# 2. Compute mean yield per YEAR
#    Which year had higher average yield?
tapply(X = barley$yield, INDEX = ___, FUN = mean)
```

```{r ex-tapply-2-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url, stringsAsFactors = TRUE)
tapply(X = barley$yield, INDEX = barley$year, FUN = mean)
```

---

## A note on the tidyverse and data.table

The functions above use base R. In practice, many R users use the
**tidyverse** (`dplyr`, `tidyr`) or **data.table** for data manipulation —
both offer more readable and often faster alternatives. We will cover these
later in the course.

For now, the base R `*apply` family is worth knowing because:
- It requires no additional packages
- The concepts transfer directly to understanding tidyverse operations
- It appears everywhere in existing R code

## Final quiz

```{r quiz-apply-final, echo=FALSE}
quiz(
  question("Which function applies a function over each group defined by a factor?",
    answer("`tapply()`", correct = TRUE),
    answer("`lapply()`"),
    answer("`apply()`"),
    answer("`mapply()`"),
    allow_retry = TRUE
  ),
  question("In `apply(X, MARGIN, FUN)`, what does `MARGIN = 2` mean?",
    answer("Apply over columns", correct = TRUE),
    answer("Apply over rows"),
    answer("Apply over the 2nd element only"),
    answer("Apply twice"),
    allow_retry = TRUE
  ),
  question("Which apply function is best for computing a summary statistic across multiple columns?",
    answer("`sapply()` or `lapply()`", correct = TRUE),
    answer("`tapply()`"),
    answer("`apply()` with MARGIN = 1"),
    answer("A for loop"),
    allow_retry = TRUE
  )
)
```
