---
title: "Functions in R"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 60)
```

## What is a function?

We have already encountered many functions in this course. A function takes
**input** (called arguments) and produces **output**. Functions allow us to
avoid repeating ourselves and to build reliable, reusable code.

## Calling ordinary functions

```{r demo-mean, echo=TRUE}
# make some dummy data
data <- c(12, 30, 2, 34, 9, 8, 64, 12, 3, 99)

# take the mean
mean(data)
```

Functions can have **parameters** — extra arguments that modify their behaviour.
The `na.rm` parameter in `mean()` is a good example:

```{r demo-na, echo=TRUE}
# append an NA (missing value)
na_data <- c(data, NA)

# mean with NA returns NA by default!
mean(na_data)
```

```{r demo-na-rm, echo=TRUE}
# pass na.rm = TRUE to remove missing values
mean(na_data, na.rm = TRUE)
```

We can also **compose** (nest) function calls:

```{r demo-compose, echo=TRUE}
data_2 <- c(102, 3, 2, 4, 9, 2, 1, 12, 97)

# sum of the two means
sum(mean(data), mean(data_2))
```

### Your turn: calling functions

```{r ex-functions-1, exercise=TRUE, exercise.lines=8}
# A biologist measured body masses (kg) of 5 penguins:
masses <- c(3.1, 4.2, NA, 3.8, 5.0)

# 1. Compute the mean — what happens with the NA?
mean(masses)

# 2. Now remove the NA and compute the mean
mean(masses, na.rm = ___)

# 3. What is the standard deviation? (function: sd)
sd(masses, na.rm = ___)
```

```{r ex-functions-1-solution}
masses <- c(3.1, 4.2, NA, 3.8, 5.0)
mean(masses)
mean(masses, na.rm = TRUE)
sd(masses, na.rm = TRUE)
```

```{r ex-compose, exercise=TRUE, exercise.lines=8}
# Compose functions: compute the range of masses (max - min)
masses <- c(3.1, 4.2, 3.8, 5.0, 4.5)

# Compute max(masses) - min(masses) in one expression
range_val <- ___(masses) - ___(masses)
range_val

# Or use the range() function and diff()
diff(range(masses))
```

```{r ex-compose-solution}
masses <- c(3.1, 4.2, 3.8, 5.0, 4.5)
range_val <- max(masses) - min(masses)
range_val
diff(range(masses))
```

```{r quiz-functions-1, echo=FALSE}
quiz(
  question("What does `na.rm = TRUE` do in `mean()`?",
    answer("Removes NA values before computing the mean", correct = TRUE),
    answer("Replaces NAs with 0"),
    answer("Returns NA if any NAs are present"),
    answer("Removes the mean from the data"),
    allow_retry = TRUE
  ),
  question("What does `sum(mean(a), mean(b))` compute?",
    answer("The sum of the mean of a and the mean of b", correct = TRUE),
    answer("The mean of all values in a and b combined"),
    answer("The sum of all values in a and b"),
    answer("An error"),
    allow_retry = TRUE
  )
)
```

---

## Special functions

R has many functions you may not think of as functions. One is `+`:

```{r demo-plus, echo=TRUE}
# normal usage
two <- 1 + 1

# captured as a function with backticks
also_two <- `+`(1, 1)

all.equal(two, also_two)
```

Another is `[`:

```{r demo-bracket, echo=TRUE}
first      <- data[1]
also_first <- `[`(data, 1)

all.equal(first, also_first)
```

You wouldn't write code like this in practice — it's terrible style — but it
shows that R's syntax is built on functions all the way down.

```{r quiz-special, echo=FALSE}
quiz(
  question("Why can `+` be called as a function in R?",
    answer("Because `+` IS a function in R — all operators are", correct = TRUE),
    answer("Because R converts it automatically"),
    answer("Because of a special language extension"),
    answer("It cannot — that code will error"),
    allow_retry = TRUE
  )
)
```

## Useful built-in functions to know

Here are some functions that will come up constantly:

```{r demo-useful, echo=TRUE}
x <- c(5, 2, 8, 1, 9, 3, 7)

length(x)   # number of elements
sum(x)      # sum all elements
cumsum(x)   # cumulative sum
sort(x)     # sort ascending
rev(x)      # reverse order
unique(c(1, 2, 2, 3, 3, 3))  # unique values
table(c("A","B","A","C","B","B"))  # frequency table
```

### Your turn: practice useful functions

```{r ex-useful, exercise=TRUE, exercise.lines=12}
# DNA sequences from 6 samples
gc_content <- c(0.48, 0.52, 0.48, 0.61, 0.55, 0.52)

# 1. How many samples?
___(gc_content)

# 2. What is the mean GC content?
___(gc_content)

# 3. Sort from lowest to highest
___(gc_content)

# 4. How many unique GC values are there?
length(unique(___))
```

```{r ex-useful-solution}
gc_content <- c(0.48, 0.52, 0.48, 0.61, 0.55, 0.52)
length(gc_content)
mean(gc_content)
sort(gc_content)
length(unique(gc_content))
```
