# Using functions on data frames

Given that we have already looked at indexing in data frames, we have already used functions on data frames! Here we will look at other functions to apply on data frames. For example:

-   How do I compute the a function on a column?

-   How do I compute the same function across several columns at once?

-   How do I compute across rows?

-   If the data is grouped, how do I execute functions on each group?

These are roughly ordered in increasing complexity.

## Functions on columns in a data frame

We have already covered this - but as a reminder.

```{r}
# load the cats data
cats <- read.csv("https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/cats.csv", stringsAsFactors = TRUE)
```

## Functions on multiple columns in a data frame

Sometimes we want to summarise several columns at the same time. It would be tedious to try and do this by hand for each one, and it would introduce errors. Less code `<-` better.

```{r}
# what are the columns to extract?
bwt_and_hwt <- c("Bwt", "Hwt")
# index on the column names
bwt_and_hwt_data <- cats[, bwt_and_hwt]

# now the fun bit
lapply(X = bwt_and_hwt_data, FUN = mean)
```

So what does this mean? `lapply` stands for 'list apply', or apply a function over elements of a list. Because we know that data frames are lists, `lapply` will treat each column as a separate list, and iterate over them. It also returns the result as a list. You also have to supply the function to `lapply` - above we gave the `mean` function. But we can very well supply our own. What if we wanted both the mean and the standard deviation?

```{r}
# bit more complex
lapply(
  X = bwt_and_hwt_data,
  # apply the functions mean and sd, but put them in a
  # list
  FUN = function(e) list(mean = mean(e), sd = sd(e))
)
```

The output is a bit of a complicated nested list. It's usually easier to use `sapply` which is a wrapper around `lapply` that simplifies the output data structure. Here it makes a little `matrix`.

```{r}
sapply(
  X = bwt_and_hwt_data,
  # apply the functions mean and sd, but put them in a
  # list
  FUN = function(e) list(mean = mean(e), sd = sd(e))
)
```

## Functions on rows in a data frame

In R, generally things are set up to compute over columns. It's a much easier and more common use case. You can apply functions on rows though, using `apply`.

```{r}
# correct for units
bwt_and_hwt_data$Bwt <- bwt_and_hwt_data$Bwt * 1000

# use apply, MARGIN = 1 means operate over rows
sum_bwt_hwt <- apply(X = bwt_and_hwt_data, MARGIN = 1, FUN = sum)

bwt_and_hwt_data$sum <- sum_bwt_hwt

head(bwt_and_hwt_data)
```

The use of `apply` is a bit overkill here. We can just use the columns. `apply` is more useful when there are many columns that you want to run a function over.

```{r}
# The above is equivalent to
bwt_and_hwt_data$sum <- bwt_and_hwt_data$Bwt + bwt_and_hwt_data$Hwt

# however over more than two columns... it can be useful
# pass other arguments to the FUN, just as comma separated arguments
# at the end.
apply(X = cats, MARGIN = 1, FUN = paste, collapse = ",")[1]
```

## Functions over groups

The `cats` data frame only has one group - sex. So we will use this simple two level factor as our grouping variable. We can look at the average body weight across the sexes using `tapply`.

```{r}
sex <- cats$Sex
body_weight <- cats$Bwt

# use tapply
tapply(X = body_weight, INDEX = sex, FUN = mean)
```

`sex` here will be forced into a `factor` which is R's internal representation of a collection of known items. Every factor will have a number of 'levels' which are the values a factor can take.

## A note on (probably) better packages

The above uses the most basic functionality in R to achieve our results. Even a cursory search of the questions above will lead to a wide variety of answers - many of them from the 'tidyverse' or 'data.table'. Each has their own syntax and quirks. We will delve later on in the course specifically to these packages to see how they work.
