# Writing functions

Functions have three things:

1. The body - the actual code inside
2. The formals - the list of arguments given to the function
3. The environment - the location of all of the functions variables

We will talk about the body and the formals, but probably not the environment until
later - that's a more complex subject.

## How to write a function

All functions are initiated with the `function` keyword. You can have functions
on single lines - I don't recommend this syntax usually, but for very short 
functions it is fine.

```{r}
# make a function which adds one to the input
add_one <- function(x) x + 1

add_one(100)
```

If your function is on more than one line, you will need curly braces to 'wrap'
the whole function nicely.

```{r}
add_one_multiply_2 <- function(x) {
  # add one
  add_one <- x + 1
  # multiply by two
  # the last line of the function is the return value
  add_one * 2
}

add_one_multiply_2(100)
```

Functions don't have to have any arguments.

```{r}
# create a function
cat_dir <- function() {
  # list.files() is a handy function
  # to list all the files in the directory you are in
  files <- list.files(".")
  
  # loop over the files - it's a vector of file names
  for (file in files) {
    # print them to the terminal
    # and put a newline at the end
    cat(file, "\n")
  }
}

# call our function
cat_dir()
```
On the other hand, functions can have default values in the arguments. See this 
example.

```{r}
# create function - then pass 'key/value' pairs in arguments
hello_name <- function(name = "Max") {
  # paste is useful for merging strings
  # there's a whole section later on string manipulation
  paste("Hello,", name)
}

# the default is to say hello to me
hi_max <- hello_name()
# but if I have to, I can say hello to Jason
hi_jason <- hello_name(name = "Jason")

# return the outputs
hi_max
hi_jason
```

