# Writing functions

Functions have three things:

1.  The body - the actual code inside
2.  The formals - the list of arguments given to the function
3.  The environment - the location of all of the functions variables

We will talk about the body and the formals, but probably not the environment until later - that's a more complex subject.

## How to write a function

All functions are initiated with the `function` keyword. You can have functions on single lines - I don't recommend this syntax usually, but for very short functions it is fine.

```{r}
# make a function which adds one to the input
add_one <- function(x) x + 1

add_one(100)
```

If your function is on more than one line, you will need curly braces to 'wrap' the whole function nicely.

```{r}
add_one_multiply_2 <- function(x) {
  # add one
  add_one <- x + 1
  # multiply by two
  # the last line of the function is the return value
  add_one * 2
}

add_one_multiply_2(100)
```

Functions don't have to have any arguments.

```{r}
# create a function
cat_dir <- function() {
  # list.files() is a handy function
  # to list all the files in the directory you are in
  files <- list.files(".")
  
  # loop over the files - it's a vector of file names
  for (file in files) {
    # print them to the terminal
    # and put a newline at the end
    cat(file, "\n")
  }
}

# call our function
cat_dir()
```

On the other hand, functions can have default values in the arguments. See this example.

```{r}
# create function - then pass 'key/value' pairs in arguments
hello_name <- function(name = "Max") {
  # paste is useful for merging strings
  # there's a whole section later on string manipulation
  paste("Hello,", name)
}

# the default is to say hello to me
hi_max <- hello_name()
# but if I have to, I can say hello to Jason
hi_jason <- hello_name(name = "Jason")

# return the outputs
hi_max
hi_jason
```

## Control flow

You can't get very far in programming without controlling the flow of information through the program. In programming terms these are the conditional statements. The `if... else` clauses. I've left this until now because control flow is most often used in functions. The syntax in general is:

```{r}
# our fake condition evaluates to true
CONDITION <- TRUE

# so we go down the true branch here
if (CONDITION) {
  TRUE
} else {
  # not the else branch!
  FALSE
}
```

The syntax is very similar to other C style programming languages.

```{r}
# if statements are useful to test input parameters

any_Rmd_files <- function(file_path) {
  # here we are testing if the file path parameter is missing
  # if it is
  if (missing(file_path)) {
    # we give a warning
    warning("You did not input a file list, defaulting to `.`\n")
    # and assing our new file path
    file_path <- "."
  }
  
  # now we can use the file path, if it was missing, it was
  # edited
  files <- list.files(file_path)
  
  # and these are our indexing rules we covered
  # we logically index into our files, if they end with Rmd
  # we keep them
  files[endsWith(files, "Rmd")]
}

any_Rmd_files()
```

`if` statements therefore don't have to have an `else` branch. But in many cases it makes sense to do so. As with functions, `if...else` statements can be on a single line. Sometimes it's easier to reason with this way.

```{r}
# we can make a quick and dirty test for a number being even
is_even <- function(x) {
  # the modulo function (%%) gives the remainder after division.
  if (x %% 2 == 0) "even" else "odd"
}

is_even(3)
```

For the last bit on control flow, we can have a look at the `ifelse` function. It's rather useful.

```{r}
# begin by loading barley
barley <- read.csv("../2_working_with_dataframes/data/barley.csv")

# take the yield column again
yield <- barley$yield

# assign high and low yield based on mean value
high_or_low_yield <- ifelse(yield < mean(yield), yes = "low", no = "high")

# assign this back to our data in a new column
barley$high_or_low_yield <- high_or_low_yield

# check the first few rows
head(barley)
```
