---
title: "Writing Functions"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 60)

barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)
```

## Anatomy of a function

Functions have three things:

1. The **body** — the actual code inside
2. The **formals** — the list of arguments given to the function
3. The **environment** — the location of all the function's variables

A function takes an optional input and produces a visible output.

## How to write a function

All functions are initiated with the `function` keyword. Short functions can go
on a single line:

```{r demo-single-line, echo=TRUE}
add_one <- function(x) x + 1
add_one(100)
```

Multi-line functions need curly braces. The **last expression** in the body is
automatically returned:

```{r demo-multi-line, echo=TRUE}
add_one_multiply_2 <- function(x) {
  add_one   <- x + 1
  add_one * 2  # this is the return value
}

add_one_multiply_2(100)
```

### Your turn: write a simple function

```{r ex-simple-fn, exercise=TRUE, exercise.lines=10}
# Write a function called `gc_percent` that:
# 1. Takes a GC content as a decimal (e.g. 0.52)
# 2. Returns it multiplied by 100 (i.e. as a percentage)

gc_percent <- function(x) {
  ___
}

gc_percent(0.52)  # should return 52
gc_percent(0.35)  # should return 35
```

```{r ex-simple-fn-solution}
gc_percent <- function(x) {
  x * 100
}

gc_percent(0.52)
gc_percent(0.35)
```

### Your turn: a multi-step function

```{r ex-multi-fn, exercise=TRUE, exercise.lines=14}
# Write a function called `bmi` that computes Body Mass Index
# BMI = mass_kg / height_m^2
# It should accept two arguments: mass_kg and height_m

bmi <- function(mass_kg, height_m) {
  ___
}

bmi(70, 1.75)   # ~22.9
bmi(90, 1.80)   # ~27.8
```

```{r ex-multi-fn-hint}
# BMI formula: mass / height^2
# Use ^ or ** for powers in R
```

```{r ex-multi-fn-solution}
bmi <- function(mass_kg, height_m) {
  mass_kg / height_m^2
}

bmi(70, 1.75)
bmi(90, 1.80)
```

---

## Default parameter values

Functions can have default values. These are used when the caller does not
supply that argument:

```{r demo-defaults, echo=TRUE}
hello_name <- function(name = "World") {
  paste("Hello,", name)
}

hello_name()         # uses default
hello_name("Jason")  # overrides default
```

### Your turn: defaults

```{r ex-defaults, exercise=TRUE, exercise.lines=14}
# Write a function called `summarise_gc` that:
# - Takes a numeric vector `gc_values`
# - Takes a `digits` argument defaulting to 2
# - Returns the mean GC content rounded to `digits` decimal places

summarise_gc <- function(gc_values, digits = ___) {
  round(mean(gc_values), digits = ___)
}

samples <- c(0.481, 0.523, 0.609, 0.547)
summarise_gc(samples)         # uses default: 2 digits
summarise_gc(samples, digits = 4)  # 4 decimal places
```

```{r ex-defaults-solution}
summarise_gc <- function(gc_values, digits = 2) {
  round(mean(gc_values), digits = digits)
}

samples <- c(0.481, 0.523, 0.609, 0.547)
summarise_gc(samples)
summarise_gc(samples, digits = 4)
```

---

## Control flow: if / else

You can't get very far in programming without controlling the flow of
information. The `if...else` clause is the main tool.

```{r demo-if-else, echo=TRUE}
CONDITION <- TRUE

if (CONDITION) {
  "went down the TRUE branch"
} else {
  "went down the FALSE branch"
}
```

A practical example — testing whether an argument is missing:

```{r demo-missing, echo=TRUE}
any_Rmd_files <- function(file_path) {
  if (missing(file_path)) {
    warning("No path given — defaulting to current directory\n")
    file_path <- "."
  }
  files <- list.files(file_path)
  files[endsWith(files, "Rmd")]
}

any_Rmd_files()
```

Single-line `if...else`:

```{r demo-even-odd, echo=TRUE}
is_even <- function(x) {
  if (x %% 2 == 0) "even" else "odd"
}

is_even(3)
is_even(8)
```

### Your turn: if / else

```{r ex-ifelse, exercise=TRUE, exercise.lines=14}
# Write a function called `classify_gc` that:
# - Takes a GC content decimal value
# - Returns "low" if < 0.4, "high" if > 0.6, "normal" otherwise

classify_gc <- function(gc) {
  if (gc < ___) {
    "low"
  } else if (gc > ___) {
    "high"
  } else {
    ___
  }
}

classify_gc(0.35)   # low
classify_gc(0.50)   # normal
classify_gc(0.68)   # high
```

```{r ex-ifelse-solution}
classify_gc <- function(gc) {
  if (gc < 0.4) {
    "low"
  } else if (gc > 0.6) {
    "high"
  } else {
    "normal"
  }
}

classify_gc(0.35)
classify_gc(0.50)
classify_gc(0.68)
```

---

## Vectorised `ifelse()`

`ifelse()` applies a condition element-wise across a whole vector:

```{r demo-ifelse-vec, echo=TRUE}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)

yield <- barley$yield

# label each row as high or low yield
high_or_low <- ifelse(yield < mean(yield), yes = "low", no = "high")

barley$yield_category <- high_or_low
head(barley)
```

### Your turn: vectorised ifelse

```{r ex-vec-ifelse, exercise=TRUE, exercise.lines=12}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)

# Use ifelse() to create a new column `era`:
# "early" for year 1931, "late" for 1932
barley$era <- ifelse(barley$year == ___, yes = "early", no = "late")

# Check the result
table(barley$era)
head(barley[, c("year", "era")])
```

```{r ex-vec-ifelse-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)
barley$era <- ifelse(barley$year == 1931, yes = "early", no = "late")
table(barley$era)
head(barley[, c("year", "era")])
```

## Final quiz

```{r quiz-writing-fns, echo=FALSE}
quiz(
  question("What is returned by a function by default?",
    answer("The result of the last expression in the function body", correct = TRUE),
    answer("NULL"),
    answer("The first expression"),
    answer("Nothing — you must use `return()`"),
    allow_retry = TRUE
  ),
  question("What does `missing(x)` return inside a function?",
    answer("TRUE if the caller did not supply `x`", correct = TRUE),
    answer("TRUE if `x` is NA"),
    answer("TRUE if `x` is NULL"),
    answer("The value of `x`"),
    allow_retry = TRUE
  ),
  question("What is the difference between `if()` and `ifelse()`?",
    answer("`if()` works on a single value; `ifelse()` is vectorised (works on whole vectors)",
           correct = TRUE),
    answer("`ifelse()` is faster but otherwise identical"),
    answer("`if()` is vectorised; `ifelse()` is not"),
    answer("There is no difference"),
    allow_retry = TRUE
  )
)
```
