---
title: "Which Statistical Test?"
author: "Max Carter-Brown"
date: "2026-03-19"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document is a **quick-reference guide** — keep it open while working through the other modules and when completing your assessed portfolio.

# The decision framework

Choosing the right test requires answering four questions in order:

1.  **What type is my response (dependent) variable?** Continuous, count, binary, ordinal?
2.  **How many groups am I comparing?** (or: is this a relationship between two continuous variables?)
3.  **Are the groups paired?** (same individuals measured twice, or matched pairs?)
4.  **Is the data normally distributed?**

# Decision table

| Response variable | Groups | Paired? | Normal? | Test | Effect size |
|----|----|----|----|----|----|
| Continuous | 1 group vs. fixed value | No | Yes | One-sample t-test | Cohen's d |
| Continuous | 2 groups | No | Yes | Independent t-test | Cohen's d |
| Continuous | 2 groups | No | No | Mann-Whitney U | Rank-biserial r |
| Continuous | 2 groups | Yes | Yes | Paired t-test | Cohen's d |
| Continuous | 2 groups | Yes | No | Wilcoxon signed-rank | Rank-biserial r |
| Continuous | 3+ groups | — | Yes | ANOVA | η² (eta-squared) |
| Continuous | 3+ groups | — | No | Kruskal-Wallis | η² |
| Continuous | Two continuous variables | — | Both normal | Pearson's correlation | r, r² |
| Continuous | Two continuous variables | — | One/both non-normal | Spearman's correlation | ρ, ρ² |
| Continuous | Predict Y from X(s) | — | Continuous Y | Linear regression / Gaussian GLM | R², pseudo-R² |
| Count | Predict Y from X(s) | — | Count Y | Poisson GLM | Pseudo-R² |
| Binary | Two categorical variables | — | — | Chi-squared / Fisher's exact | Cramér's V |
| Binary | Predict Y from X(s) | — | Binary Y | Binomial GLM (logistic regression) | Pseudo-R² |
| Any | Multiple studies pooled | — | — | Meta-analysis | Pooled SMD, I² |

# Decision flowchart

```         
What is your response (Y) variable?
│
├── CATEGORICAL (binary, nominal, ordinal)
│   └── Are you comparing observed vs. expected counts?
│       ├── Yes → Chi-squared test (or Fisher's if cells < 5)
│       │         Effect size: Cramér's V
│       └── No  → Chi-squared / Fisher's exact for two categorical variables
│
├── CONTINUOUS (or count data)
│   └── Are you asking about RELATIONSHIPS or GROUP DIFFERENCES?
│       │
│       ├── RELATIONSHIP (two continuous variables)
│       │   └── Both normal? → Pearson's r
│       │       One/both non-normal? → Spearman's ρ
│       │       Want to predict Y from X? → Linear regression
│       │
│       └── GROUP DIFFERENCES
│           └── How many groups?
│               │
│               ├── 2 groups
│               │   └── Paired measurements?
│               │       ├── Yes → Paired t-test (normal) / Wilcoxon signed-rank (non-normal)
│               │       └── No  → t-test (normal) / Mann-Whitney U (non-normal)
│               │
│               └── 3+ groups
│                   └── ANOVA (normal) / Kruskal-Wallis (non-normal)
│
└── COUNT DATA
    └── Poisson GLM
```

# Checking normality

Always check normality *before* choosing parametric vs. non-parametric tests. Use both:

1.  **Visual check**: histogram with binwidth ≈ SD/3. Does it look bell-shaped?
2.  **Shapiro-Wilk test**: `shapiro.test()`. p \> 0.05 = consistent with normality.

> **Important:** Shapiro-Wilk has low power with small samples (n \< 30). A non-significant result with n = 15 does *not* confirm normality. Rely more on the histogram with small samples.

```{r normality-examples}
library(ggplot2)
set.seed(42)

normal_data    <- rnorm(100, mean = 50, sd = 10)
skewed_data    <- rexp(100, rate = 0.1)
bimodal_data   <- c(rnorm(50, mean = 40, sd = 5), rnorm(50, mean = 60, sd = 5))

par(mfrow = c(1, 3))
hist(normal_data,  main = "Normal",  xlab = "Value", col = "steelblue", breaks = 15)
hist(skewed_data,  main = "Skewed",  xlab = "Value", col = "tomato",    breaks = 15)
hist(bimodal_data, main = "Bimodal", xlab = "Value", col = "darkgreen", breaks = 15)
par(mfrow = c(1, 1))

cat("Shapiro-Wilk:\n")
cat("Normal  p =", round(shapiro.test(normal_data)$p.value,  4), "\n")
cat("Skewed  p =", round(shapiro.test(skewed_data)$p.value,  4), "\n")
cat("Bimodal p =", round(shapiro.test(bimodal_data)$p.value, 4), "\n")
```

# Effect sizes — quick reference

Always report an effect size. A p-value tells you whether an effect exists; an effect size tells you how large it is.

| Test | Effect size | Interpretation |
|----|----|----|
| t-test | Cohen's d | 0.2 small, 0.5 medium, 0.8 large |
| Mann-Whitney | Rank-biserial r | 0.1 small, 0.3 medium, 0.5 large |
| ANOVA / Kruskal-Wallis | η² (eta-squared) | 0.01 small, 0.06 medium, 0.14 large |
| Chi-squared | Cramér's V | \< 0.1 negligible, 0.1–0.3 weak, 0.3–0.5 moderate, \> 0.5 strong |
| Pearson/Spearman | r (correlation) | 0.1 weak, 0.3 moderate, 0.5 strong |
| Regression | R² | proportion of variance explained |
| GLM | McFadden's pseudo-R² | proportion of deviance explained |
| Meta-analysis | Pooled SMD | 0.2 small, 0.5 medium, 0.8 large |

# Reporting p-values

-   **p \< 0.001**: report as p \< 0.001 (not p = 0.000)
-   **p = 0.042**: report to 3 decimal places
-   **p \> 0.05**: report as p = \_\_\_ (not "ns" alone); include the actual value
-   Always accompany with test statistic, degrees of freedom, and effect size

**Template (t-test):** \> *"Males were significantly taller than females (t(df) = value, p = value, Cohen's d = value)."*

**Template (chi-squared):** \> *"There was a significant association between sex and smoking status (χ²(df) = value, p = value, Cramér's V = value)."*

**Template (correlation):** \> *"Height was significantly positively correlated with weight (Pearson's r(df) = value, p = value)."*

**Template (regression):** \> *"Height significantly predicted weight (β = value, F(df1, df2) = value, p = value, R² = value)."*

------------------------------------------------------------------------

# Quick practice

```{r load-data}
d <- read.csv(
  "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/statistics/data/form-1__anthropometrics.csv",
  header = TRUE, stringsAsFactors = TRUE
)
```

For each question below, (a) state which test to use and why, (b) run the test, (c) report the result in a sentence.

## Question 1

Does body temperature (`X13_Body_temp_celsius`) differ between smokers and non-smokers (`X8_Smoker`)?

*Insert R code and answer here*

## Question 2

Is there a relationship between age (`X4_Age_in_years`) and systolic blood pressure (`X14_Systolic_blood_pre`)?

*Insert R code and answer here*

## Question 3

Is blood group (`X9_Blood_group`) independent of sex (`X3_Gender`)?

*Insert R code and answer here*

## Question 4

Does activity level (`X7_Activity_Level_NAS`) affect BMI? First create a BMI column.

*Insert R code and answer here*
