---
title: "Study Design and the Role of Statistics"
author: "Max Carter-Brown"
date: "2026-03-19"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(ggplot2)
```

# Why does study design matter?

Statistics cannot rescue a badly designed study. The choices made *before* you collect any data — what to measure, how many samples to take, how to assign treatments — determine what conclusions you can validly draw. This module introduces the language and principles of good study design so that the statistical tests in subsequent modules make sense in context.

> **Common misconception:** If a statistical test gives p \< 0.05, the result is meaningful. A significant p-value only tells you the result is unlikely under the null hypothesis. A biased or underpowered study can produce significant results that are meaningless, unrepeatable, or both.

# Hypotheses

Every study begins with a question, which becomes a **hypothesis** — a testable, falsifiable statement about the world.

-   **Null hypothesis (H₀)**: No effect, no difference, no relationship. The default assumption.
-   **Alternative hypothesis (H₁)**: Something is going on. **What you expect to find**.

**Example:** H₀: Mean height does not differ between males and females. H₁: Mean height differs between males and females.

You never *prove* H₁. You accumulate evidence *against* H₀ until it becomes implausible.

## One-tailed vs two-tailed tests

A **two-tailed** test asks "is there any difference?" (H₁: μ₁ ≠ μ₂). A **one-tailed** test asks "is group A specifically larger/smaller?" (H₁: μ₁ \> μ₂). One-tailed tests have more power but require a strong directional prediction made *before* looking at the data. Using a one-tailed test to rescue a borderline two-tailed result is **p-hacking**.

# Types of variables

| Type | Description | Example | Appropriate test |
|---------------|------------------|---------------|------------------------|
| **Continuous** | Any real number | Height, weight, temperature | t-test, ANOVA, regression |
| **Count** | Non-negative integers | Heart rate (bpm), cell counts | Poisson GLM |
| **Binary** | Two outcomes | Smoker (yes/no) | Chi-squared, logistic regression |
| **Ordinal** | Ordered categories | Activity level (low/medium/high) | Mann-Whitney, Kruskal-Wallis, advanced (ordinal) GLM |
| **Nominal** | Unordered categories | Blood group (A/B/AB/O) | Chi-squared, advanced (multinomial) GLM |

Identifying your variable types *before* analysis determines your statistical approach.

# Independent vs dependent variables

-   **Independent variable (predictor, X)**: What you control or measure as a potential cause.
-   **Dependent variable (response, Y)**: What you measure as a potential effect.

"Height predicts weight" → height = X, weight = Y.

Keep this distinction clear. Regression and ANOVA are asymmetric: swapping X and Y gives a different model.

# Types of study

## Observational studies

You measure variables as they are — no manipulation. Examples: surveys, cohort studies, case-control studies.

-   **Advantage**: Can study relationships in natural populations.
-   **Limitation**: Cannot establish causation. Confounders may explain any observed association.

## Experimental studies

You **randomly assign** participants to treatment and control groups.

-   **Advantage**: Randomisation balances confounders across groups, allowing causal inference.
-   **Limitation**: Often impractical, expensive, or unethical for many biological questions.

> "Correlation does not imply causation" — but randomised experiments can.

# Confounding variables

A **confounder** is a variable that is associated with both the independent and dependent variable, creating a spurious apparent relationship.

```{r confounding-simulation}
set.seed(42)
n <- 200

# True causal structure: age → height AND age → weight
# There is no direct causal path from height to weight here
age    <- rnorm(n, mean = 20, sd = 3)
height <- 150 + 0.5 * age + rnorm(n, sd = 8)
weight <- 40  + 0.8 * age + rnorm(n, sd = 10)

cor_hw <- round(cor(height, weight), 2)
cat("Correlation between height and weight:", cor_hw, "\n")
cat("(Even though neither directly causes the other — age causes both)\n")

cor_aw <- round(cor(age, weight), 2)
cat("Stronger correlation between age and weight:", cor_aw, "\n")
```

The lesson: a strong correlation between height and weight here is entirely driven by age. Without measuring age, you would wrongly conclude height and weight are directly related.

# Sample size and statistical power

**Power** is the probability of detecting a real effect when one exists. Low power leads to: - Failing to detect real effects (**Type II errors**, false negatives) - Inflated effect size estimates in the studies that *do* detect effects (the "winner's curse")

Power depends on three things:

| Factor        | Effect on power |
|---------------|-----------------|
| Sample size ↑ | Power ↑         |
| Effect size ↑ | Power ↑         |
| Variability ↑ | Power ↓         |

```{r power-simulation}
# Simulate detecting a mean difference of d standard deviations
set.seed(42)

simulate_power <- function(n, delta, n_sims = 500) {
  # delta = true difference in means (in SDs)
  p_vals <- replicate(n_sims, {
    a <- rnorm(n, mean = 0,     sd = 1)
    b <- rnorm(n, mean = delta, sd = 1)
    t.test(a, b)$p.value
  })
  mean(p_vals < 0.05)
}

results <- expand.grid(n = c(10, 30, 50, 100), delta = c(0.2, 0.5, 0.8))
results$power <- mapply(simulate_power, results$n, results$delta)
results$delta <- factor(results$delta, labels = c("Small (d=0.2)", "Medium (d=0.5)", "Large (d=0.8)"))

ggplot(results, aes(x = n, y = power, colour = delta, group = delta)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("tomato", "steelblue", "darkgreen")) +
  xlab("Sample size per group") + ylab("Power (probability of detecting the effect)") +
  labs(colour = "Effect size", caption = "Dashed line = 80% power (conventional minimum)") +
  theme_bw()
```

For a medium effect, you need roughly **n = 50 per group** to reach 80% power. Small effects require much larger samples. This is why a non-significant result from a small study is almost uninterpretable — the study may simply have been too small to detect the effect. The more (random) samples, the better!

# The p-value: what it is and what it is not

The p-value is the probability of observing data as extreme as yours (or more extreme) **if the null hypothesis were true**.

| The p-value IS                     | The p-value IS NOT         |
|------------------------------------|----------------------------|
| A measure of compatibility with H₀ | The probability H₀ is true |
| Affected by sample size            | A measure of effect size   |
| One piece of evidence              | A binary verdict on truth  |

Always report effect sizes and confidence intervals alongside p-values.

------------------------------------------------------------------------

# Exercises

## Question 1. Identify variable types

For each variable in the class data, identify whether it is continuous, count, binary, ordinal, or nominal.

```{r load-data}
d <- read.csv(
  "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/statistics/data/form-1__anthropometrics.csv",
  header = TRUE, stringsAsFactors = TRUE
)
names(d)
```

*Write your answers here*

## Question 2. Formulate hypotheses

Write null and alternative hypotheses (H₀ and H₁) for the following research questions:

a.  Does body temperature differ between smokers and non-smokers?
b.  Is BMI associated with systolic blood pressure?
c.  Does activity level affect heart rate?

*Write your answers here*

## Question 3. Identify the independent and dependent variables in each question above.

*Write your answers here*

## Question 4. Confounding

For question (a) above — does sex confound the relationship between smoking and body temperature? Explain in one sentence why or why not.

*Write your answer here*

## Challenge Question 5. Power calculation

Using the simulation code above, estimate what sample size per group you would need to have 80% power to detect a small effect (d = 0.2). Report your answer and explain your reasoning.

*Insert R code and interpretation here*
