---
title: "Reading Error Messages"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 60)
```

## Why errors are your friends

Every programmer — beginner and expert alike — encounters errors constantly.
The key skill is being able to **read** an error message and understand what it
is telling you. R's errors are often very direct once you know how to read them.

This module covers the most common errors you will encounter, what they mean,
and how to fix them.

> **Rule of thumb:** Read the *last line* of the error first. That's usually
> the most informative part.

---

## Error vs Warning vs Message

R has three levels of feedback:

| Type | What it means |
|------|--------------|
| **Error** | Something went wrong; execution stopped |
| **Warning** | Something suspicious happened; execution continued |
| **Message** | Informational output; not a problem |

```{r demo-error, echo=TRUE, error=TRUE}
# This produces an ERROR - execution stops
sqrt("hello")
```

```{r demo-warning, echo=TRUE, warning=TRUE}
# This produces a WARNING - execution continues, but something is odd
log(-1)
```

```{r demo-message, echo=TRUE, message=TRUE}
# Messages are just informational (e.g. package loading)
message("This is just an informational message, not an error")
```

```{r quiz-types, echo=FALSE}
quiz(
  question("Which type of R output stops code execution?",
    answer("Error", correct = TRUE),
    answer("Warning"),
    answer("Message"),
    answer("All three"),
    allow_retry = TRUE
  )
)
```

---

## Error 1: Object not found

**Message:** `Error: object 'x' not found`

This is the most common beginner error. It means you are trying to use a
variable that doesn't exist in the current environment — usually because:
- You made a typo in the variable name
- You forgot to run the line that creates the variable
- You created it in a different script or session

```{r demo-not-found, echo=TRUE, error=TRUE}
# We never created `speices` (note the typo!)
mean(speices)
```

**Fix:** Check the spelling, and make sure the variable has been created.

### Your turn: fix the object not found error

```{r ex-not-found, exercise=TRUE, exercise.lines=6}
# This code has a typo — fix it so it runs without error
species_count <- c(12, 45, 8, 33, 21)
total <- sum(species_counts)   # <-- error here!
total
```

```{r ex-not-found-hint}
# Look carefully at the variable name in sum() vs the one you created
```

```{r ex-not-found-solution}
species_count <- c(12, 45, 8, 33, 21)
total <- sum(species_count)   # fixed: removed the 's'
total
```

---

## Error 2: Could not find function

**Message:** `Error: could not find function "myfunction"`

This means R doesn't know about the function. Usually because:
- You misspelled the function name
- You forgot to load a package that provides the function
- The function doesn't exist

```{r demo-no-fn, echo=TRUE, error=TRUE}
# calculate_mean doesn't exist — did you mean mean()?
calculate_mean(c(1, 2, 3))
```

**Fix:** Check spelling (`?mean`, `?read.csv`). If the function is from a
package, load it with `library(packagename)`.

### Your turn: fix the function name error

```{r ex-no-fn, exercise=TRUE, exercise.lines=6}
# Fix the two function name errors below
masses <- c(3.1, 4.2, 3.8, 5.0)
result <- Median(masses)           # error 1
cat("Median:", result, "\n")

sd_val <- standard_deviation(masses)  # error 2
cat("SD:", sd_val, "\n")
```

```{r ex-no-fn-hint}
# R is case-sensitive: Median() vs median()
# standard_deviation() doesn't exist — the base R function is sd()
```

```{r ex-no-fn-solution}
masses <- c(3.1, 4.2, 3.8, 5.0)
result <- median(masses)
cat("Median:", result, "\n")

sd_val <- sd(masses)
cat("SD:", sd_val, "\n")
```

---

## Error 3: Non-numeric argument to binary operator

**Message:** `Error in ... : non-numeric argument to binary operator`

You tried to do arithmetic on something that isn't a number.

```{r demo-non-numeric, echo=TRUE, error=TRUE}
weight <- "3.5 kg"   # stored as a string, not a number!
weight * 2
```

**Fix:** Make sure numeric values aren't stored as characters. Use `as.numeric()`
if needed. Check your data with `class()` or `str()`.

### Your turn: fix the type error

```{r ex-non-numeric, exercise=TRUE, exercise.lines=8}
# These values are characters because they have units embedded
# Fix the code so the mean computes correctly
heights <- c("1.65", "1.72", "1.80", "1.58")

# Convert heights to numeric first, then compute mean
mean_height <- mean(___(heights))
mean_height
```

```{r ex-non-numeric-solution}
heights <- c("1.65", "1.72", "1.80", "1.58")
mean_height <- mean(as.numeric(heights))
mean_height
```

---

## Error 4: Subscript out of bounds

**Message:** `Error: subscript out of bounds`

You tried to access an element at an index that doesn't exist.

```{r demo-bounds, echo=TRUE, error=TRUE}
x <- c(10, 20, 30)
x[10]   # there is no 10th element — returns NA, not an error for vectors...

# but for lists or matrices, it's an actual error:
my_list <- list(a = 1, b = 2)
my_list[[5]]   # no 5th element!
```

**Fix:** Check the length of your object with `length()`, `nrow()`, or `ncol()`.

### Your turn: diagnose a subscript error

```{r ex-bounds, exercise=TRUE, exercise.lines=8}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)

# How many rows does barley have?
nrow(barley)

# This line tries to get a row that may not exist — find the last valid row
# and fix the index
barley[200, ]   # does row 200 exist? Check nrow() first!
```

```{r ex-bounds-hint}
# nrow(barley) tells you the maximum valid row index
# If nrow is 120, valid indices are 1 to 120
```

```{r ex-bounds-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)
nrow(barley)   # 120 rows — so max index is 120
barley[120, ]  # last row
```

---

## Error 5: Unexpected symbol / unexpected end of input

**Message:** `Error: unexpected symbol in "..."` or `unexpected end of input`

R couldn't parse your code — usually a syntax error:
- Missing a closing bracket `)`, `}`, or `]`
- Missing a comma between arguments
- Using `=` where you meant `==`

```{r demo-syntax, echo=TRUE, error=TRUE}
# Missing closing parenthesis
# result <- mean(c(1, 2, 3)   # would error if uncommented

# Missing comma
result <- c(1 2 3)  # error: missing commas
```

### Your turn: fix syntax errors

```{r ex-syntax, exercise=TRUE, exercise.lines=8}
# Fix the three syntax errors in this code

my_data <- c(10, 20 30, 40)

if (mean(my_data > 20) {
  cat("Mean is above 20\n")
}
```

```{r ex-syntax-hint}
# 1. Missing comma between 20 and 30
# 2. The closing ) on if() is in the wrong place
# 3. mean(my_data) > 20, not mean(my_data > 20) — different logic!
```

```{r ex-syntax-solution}
my_data <- c(10, 20, 30, 40)   # added missing comma

if (mean(my_data) > 20) {      # moved ) to after my_data, > 20 outside
  cat("Mean is above 20\n")
}
```

---

## How to get help

### 1. The `?` operator

```{r demo-help, eval=FALSE}
?mean          # opens the help page for mean()
?read.csv      # help for read.csv
?"["           # help for the indexing operator
```

### 2. Google the error message

Copy the error message (without your specific variable names) and paste it into
Google. Add "R" at the start. For example:

> R "non-numeric argument to binary operator"

Stack Overflow will have an answer almost every time.

### 3. Check your variable's type

When something doesn't behave as expected, always check:

```{r demo-debug, echo=TRUE}
x <- "3.14"  # looks like a number, but it's a string!
class(x)
is.numeric(x)
```

```{r quiz-errors-final, echo=FALSE}
quiz(
  question("You get `Error: object 'df' not found`. What is the most likely cause?",
    answer("The variable `df` was never created, or was created with a different name",
           correct = TRUE),
    answer("The data frame is empty"),
    answer("`df` is a reserved keyword in R"),
    answer("You need to install a package"),
    allow_retry = TRUE
  ),
  question("You get `could not find function 'ggplot'`. What should you do?",
    answer("Run `library(ggplot2)` to load the package first", correct = TRUE),
    answer("Reinstall R"),
    answer("Use `ggplot::plot()` instead"),
    answer("Check if the data frame has a column called ggplot"),
    allow_retry = TRUE
  ),
  question("What is the best first step when you encounter an error message?",
    answer("Read the error message carefully — especially the last line", correct = TRUE),
    answer("Delete the offending line of code"),
    answer("Restart R and try again"),
    answer("Ask for help immediately"),
    allow_retry = TRUE
  ),
  question("What function lets you check the type of an object?",
    answer("`class()` or `typeof()`", correct = TRUE),
    answer("`type()`"),
    answer("`kind()`"),
    answer("`format()`"),
    allow_retry = TRUE
  )
)
```
