---
editor_options: 
  markdown: 
    wrap: 72
---

# How do I do a Poisson GLM and plot the predictions in R?

## Read in the data

First of all read in the data, and run the model. If you are unsure how
to do this, please check out the `R statistics` part of the website.

```{r}
d <- read.csv("https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/statistics/data/form-1__anthropometrics.csv")
# save the model. Here, we're modelling heart rate as a function of height of the person
mod <- glm(X19_Heart_rate_bpm_me ~ X5_Height_in_cm, family = "poisson", data = d)
```

## Summary of the model

```{r}
mod_summary <- summary(mod)

mod_summary

cat("Data scale intercept is: ", exp(mod_summary$coefficients[1, 1]), ".\n", sep = "")
cat("Data scale slope is: ", exp(mod_summary$coefficients[2, 1]), ".", sep = "")
```

Interpretation is as follows. Intercept is `exp(5.525589)` and at zero
height `exp(5.525589 - 0.006652)`. The slope is `exp(-0.006652)`. This
means that for every unit increase in height, heart rate goes down by
around 1%. We'll see this visually later. We can also work out what the
predicted heart rate would be for a person who is 160cm tall.

```{r}
# hear
at_160 <- exp(mod_summary$coefficients[1, 1] + mod_summary$coefficients[2, 1] * 160)
at_180 <- exp(mod_summary$coefficients[1, 1] + mod_summary$coefficients[2, 1] * 180)

cat("Predicted heart rate at 160cm:", at_160, "\n")
cat("And there's a", at_160 / at_180, "x higher heart rate at 160cm than at 180cm\n")
```

## Predict the data

Now we want to plot our predicted values on top of the raw values from
the data.

```{r}
# interpolate within the bounds of our data
min_height <- min(d$X5_Height_in_cm)
max_height <- max(d$X5_Height_in_cm)
# 1000 data points to simulate
new_x <- expand.grid(
  X5_Height_in_cm = seq(min_height, max_height, length = 1000)
)
# and predict the y using the `predict` function
# se.fit to get the confidence intervals
new_y <- as.data.frame(predict(mod, newdata = new_x, se.fit = TRUE))
# modify the new data frame
new_d <- data.frame(new_x, new_y)
# in particular we need to exponentiate, as poisson
# is a log link scale
new_d$fit_exp <- exp(new_d$fit) 
# 95% of the area under the normal distribution lies within 
# 1.96 standard deviations away from the mean.
new_d$lwr <- exp(new_d$fit - 1.96 * new_d$se.fit) 
new_d$upr <- exp(new_d$fit + 1.96 * new_d$se.fit)

head(new_d)
```

## Plot the data

```{r}
plot(d$X5_Height_in_cm, d$X19_Heart_rate_bpm_me, xlab = "Height (cm)", ylab = "Heart rate (bpm)")
polygon(
  x = c(new_d$X5_Height_in_cm,  rev(new_d$X5_Height_in_cm)),
  c(new_d$lwr, rev(new_d$upr)), col = alpha("red", 0.2), 
  border = NA
)

# add the lines
lines(new_d$X5_Height_in_cm, new_d$fit_exp)
lines(new_d$X5_Height_in_cm, new_d$lwr)
lines(new_d$X5_Height_in_cm, new_d$upr)
```
```{r, echo=FALSE}
# now export data
write.csv(
  x = new_d[, c("X5_Height_in_cm", "fit_exp", "lwr", "upr")],
  file = "./data/poisson.csv"
)
```

References: -
<https://stackoverflow.com/questions/75017400/plot-standard-error-in-base-r-scatterplot> -
Getting Started with R - Andrew Beckerman -
<https://stats.stackexchange.com/questions/272194/interpreting-poisson-output-in-r> -
<https://stats.stackexchange.com/questions/11096/how-to-interpret-coefficients-in-a-poisson-regression>
