---
title: "Capstone Project: Penguins of Palmer Station"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 120)

penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins_raw <- read.csv(penguins_url)
penguins     <- na.omit(penguins_raw)
```

## Welcome to the Capstone!

You have now covered all the core building blocks:

- Variables and data types
- Reading and manipulating data frames
- Writing functions
- Producing plots

In this capstone you will work through a **complete, realistic data analysis**
from start to finish using the **Palmer Penguins** dataset — morphological
measurements of 344 penguins from three species (*Adelie*, *Chinstrap*, and
*Gentoo*) on three Antarctic islands.

**Your mission:** Describe the penguins, find patterns in their morphology,
compare species, and produce a reproducible report in code.

---

## Step 1: Import and inspect the data

```{r step1-demo, echo=TRUE}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins_raw <- read.csv(penguins_url)
```

### Exercise 1.1 — Basic inspection

```{r ex-inspect, exercise=TRUE, exercise.lines=10}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins_raw <- read.csv(penguins_url)

# Answer these questions using R:
# 1. How many rows and columns?
dim(___)

# 2. What are the column names?
colnames(___)

# 3. What types are the columns?
str(___)
```

```{r ex-inspect-solution}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins_raw <- read.csv(penguins_url)
dim(penguins_raw)
colnames(penguins_raw)
str(penguins_raw)
```

### Exercise 1.2 — How many missing values?

```{r ex-missing, exercise=TRUE, exercise.lines=10}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins_raw <- read.csv(penguins_url)

# Count NAs in each column using sapply and is.na
# is.na(x) returns TRUE for each NA; sum() counts them
sapply(penguins_raw, function(col) sum(___))
```

```{r ex-missing-hint}
# is.na(col) gives TRUE/FALSE for each element
# sum(is.na(col)) counts the NAs
```

```{r ex-missing-solution}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins_raw <- read.csv(penguins_url)
sapply(penguins_raw, function(col) sum(is.na(col)))
```

---

## Step 2: Clean the data

### Exercise 2.1 — Remove missing values and convert to factors

```{r ex-clean, exercise=TRUE, exercise.lines=14}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins_raw <- read.csv(penguins_url)

# 1. Remove rows with any NA values using na.omit()
penguins <- na.omit(___)

# 2. Convert species, island, and sex to factors
penguins$species <- as.factor(penguins$___)
penguins$island  <- as.factor(penguins$___)
penguins$sex     <- as.factor(penguins$___)

# 3. Check the result
str(penguins)
```

```{r ex-clean-solution}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins_raw <- read.csv(penguins_url)
penguins <- na.omit(penguins_raw)
penguins$species <- as.factor(penguins$species)
penguins$island  <- as.factor(penguins$island)
penguins$sex     <- as.factor(penguins$sex)
str(penguins)
```

---

## Step 3: Summarise the data

### Exercise 3.1 — How many of each species?

```{r ex-count, exercise=TRUE, exercise.lines=6}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))

# Use table() to count penguins per species
table(penguins$___)
```

```{r ex-count-solution}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))
table(penguins$species)
```

### Exercise 3.2 — Mean body mass per species

```{r ex-mass-by-species, exercise=TRUE, exercise.lines=8}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))

# Use tapply to compute mean body_mass_g per species
tapply(X = penguins$___, INDEX = penguins$___, FUN = ___)
```

```{r ex-mass-by-species-solution}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))
tapply(X = penguins$body_mass_g, INDEX = penguins$species, FUN = mean)
```

### Exercise 3.3 — Multi-column summary

```{r ex-summary, exercise=TRUE, exercise.lines=12}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))

# Use sapply to get mean and SD for all four numeric measurement columns
numeric_cols <- penguins[, c("bill_length_mm", "bill_depth_mm",
                              "flipper_length_mm", "body_mass_g")]

sapply(numeric_cols, function(col) c(mean = ___(col), sd = ___(col)))
```

```{r ex-summary-solution}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))
numeric_cols <- penguins[, c("bill_length_mm", "bill_depth_mm",
                              "flipper_length_mm", "body_mass_g")]
sapply(numeric_cols, function(col) c(mean = mean(col), sd = sd(col)))
```

---

## Step 4: Filter the data

### Exercise 4.1 — Large Gentoo penguins

```{r ex-filter, exercise=TRUE, exercise.lines=10}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))

# Filter to Gentoo penguins with body_mass_g > 5500
is_gentoo    <- penguins$species ___ "Gentoo"
is_large     <- penguins$body_mass_g ___ 5500

big_gentoo <- penguins[is_gentoo ___ is_large, ]
nrow(big_gentoo)
head(big_gentoo)
```

```{r ex-filter-solution}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))
is_gentoo  <- penguins$species == "Gentoo"
is_large   <- penguins$body_mass_g > 5500
big_gentoo <- penguins[is_gentoo & is_large, ]
nrow(big_gentoo)
head(big_gentoo)
```

---

## Step 5: Write a function

### Exercise 5.1 — A body condition index function

Body Condition Index (BCI) = `body_mass_g / flipper_length_mm`

```{r ex-function, exercise=TRUE, exercise.lines=18}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))

# Write a function called `body_condition` that:
# - Takes a data frame with columns body_mass_g and flipper_length_mm
# - Returns a numeric vector of BCI values (mass / flipper length)

body_condition <- function(df) {
  df$___ / df$___
}

# Apply it to the penguins data frame
penguins$bci <- body_condition(___)

# What is the mean BCI per species?
tapply(penguins$bci, penguins$species, mean)
```

```{r ex-function-solution}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))

body_condition <- function(df) {
  df$body_mass_g / df$flipper_length_mm
}

penguins$bci <- body_condition(penguins)
tapply(penguins$bci, penguins$species, mean)
```

---

## Step 6: Visualise

### Exercise 6.1 — Distribution of bill lengths by species

```{r ex-plot-1, exercise=TRUE, exercise.lines=12}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))

# Box plot of bill_length_mm by species using ggplot2
# Fill by species, use theme_bw()
ggplot(data = penguins, aes(x = ___, y = ___, fill = ___)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Bill length by species",
    x     = ___,
    y     = "Bill length (mm)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

```{r ex-plot-1-solution}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))
ggplot(data = penguins, aes(x = species, y = bill_length_mm, fill = species)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Bill length by species",
    x     = "Species",
    y     = "Bill length (mm)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

### Exercise 6.2 — Scatter plot with species coloured

```{r ex-plot-2, exercise=TRUE, exercise.lines=14}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))

# Scatter plot: flipper_length_mm (x) vs body_mass_g (y)
# Colour by species, add a linear trend line per species
# Use facet_wrap by sex
ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g, colour = ___)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ ___) +
  labs(title = "Flipper length vs body mass by sex",
       x = "Flipper length (mm)", y = "Body mass (g)",
       colour = "Species") +
  theme_bw()
```

```{r ex-plot-2-solution}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))
ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g, colour = species)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ sex) +
  labs(title = "Flipper length vs body mass by sex",
       x = "Flipper length (mm)", y = "Body mass (g)",
       colour = "Species") +
  theme_bw()
```

---

## Step 7: Save your results

### Exercise 7.1 — Write a filtered dataset

```{r ex-write, exercise=TRUE, exercise.lines=14}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))

# Filter to only Adelie penguins
adelie <- penguins[penguins$species ___ "Adelie", ]

# Write to a temp CSV (no row names)
tmp <- tempfile(fileext = ".csv")
write.csv(x = ___, file = tmp, row.names = ___)

# Read back and confirm
result <- read.csv(tmp)
cat("Rows saved:", nrow(result), "\n")
table(result$species)
```

```{r ex-write-solution}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- na.omit(read.csv(penguins_url))
adelie <- penguins[penguins$species == "Adelie", ]
tmp <- tempfile(fileext = ".csv")
write.csv(x = adelie, file = tmp, row.names = FALSE)
result <- read.csv(tmp)
cat("Rows saved:", nrow(result), "\n")
table(result$species)
```

---

## Congratulations!

You have completed a full data analysis workflow:

1. **Imported** a real dataset from the web
2. **Inspected** its structure and identified missing values
3. **Cleaned** the data by removing NAs and fixing column types
4. **Summarised** key variables by group using `tapply` and `sapply`
5. **Filtered** the data with logical indexing
6. **Wrote a function** to compute a new biological variable
7. **Visualised** distributions and relationships with ggplot2
8. **Saved** a filtered result to disk

These are the steps of almost every real data analysis. Well done!

## Final reflection quiz

```{r quiz-capstone, echo=FALSE}
quiz(
  question("Which function removes rows with any NA values?",
    answer("`na.omit()`", correct = TRUE),
    answer("`remove.na()`"),
    answer("`filter_na()`"),
    answer("`drop_na()` (tidyverse only, not base R)",
           message = "Correct that drop_na() exists but it's tidyverse — na.omit() is base R."),
    allow_retry = TRUE
  ),
  question("You want the mean of a column, grouped by species. Which function is best?",
    answer("`tapply()`", correct = TRUE),
    answer("`sapply()`"),
    answer("`apply()` with MARGIN = 2"),
    answer("`lapply()`"),
    allow_retry = TRUE
  ),
  question("In ggplot2, how do you create separate panels per sex?",
    answer("`facet_wrap(~ sex)`", correct = TRUE),
    answer("`panel(sex)`"),
    answer("`aes(panel = sex)`"),
    answer("`split_by(sex)`"),
    allow_retry = TRUE
  )
)
```
