---
title: "Classes in R"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 60)
```

## What are classes?

Classes are extended data types either defined by the people who maintain R, or
us, the humble users. There are at least three different ways of defining classes
in R (`S3`, `S4`, `Reference classes`). We only look at `S3` here, as they are
the most simple (if a bit weird).

> **Note:** This is a relatively advanced topic. Don't worry if it doesn't fully
> click right away — revisit it once you're comfortable with functions and data frames.

## Built-in classes

A good example is `data.frame`, which is a glorified `list` with extra attributes.

```{r demo-df-class, echo=TRUE}
d <- data.frame(a = 1:10, b = 11:20, c = 21:30)

# class() tells you what class an object is
cat("Class:", class(d), "\n")
# typeof() tells you the underlying storage type
cat("Type:", typeof(d), "\n")
```

We can see all the methods associated with a class — these are functions that
behave differently depending on what object you pass to them.

```{r demo-df-methods, echo=TRUE}
# list all methods defined for data.frame
methods(class = "data.frame")
```

The `plot()` function is a good example of a **generic** function. Call it on a
`data.frame` and it tries to plot every pair of columns:

```{r demo-plot-df, echo=TRUE}
plot(d)
```

But call `plot()` on a `table` and you get a bar chart:

```{r demo-plot-table, echo=TRUE}
# sample 30 letters from a-j
s   <- sample(letters[1:10], 30, replace = TRUE)
tab <- table(s)
plot(tab)
```

Same function name (`plot`), completely different behaviour — because the class
of the input is different. This is called **polymorphism**.

### Your turn: exploring classes

```{r ex-class-explore, exercise=TRUE, exercise.lines=8}
# 1. Create a numeric vector called `lengths` with values 1 through 10
# 2. Check its class with class()
# 3. Check its type with typeof()
# 4. How many methods does the "numeric" class have? (use length(methods(...)))
lengths_vec <- 1:10
class(___)
typeof(___)
length(methods(class = ___))
```

```{r ex-class-explore-hint}
# class() and typeof() take the variable name
# methods(class = "numeric") lists all methods for numerics
```

```{r ex-class-explore-solution}
lengths_vec <- 1:10
class(lengths_vec)
typeof(lengths_vec)
length(methods(class = "numeric"))
```

```{r quiz-classes-1, echo=FALSE}
quiz(
  question("What does `class()` return for a `data.frame`?",
    answer('"data.frame"', correct = TRUE),
    answer('"list"'),
    answer('"numeric"'),
    answer('"table"'),
    allow_retry = TRUE
  ),
  question("Two objects have different classes but you call the same function on both and get different results. What is this called?",
    answer("Polymorphism", correct = TRUE),
    answer("Coercion"),
    answer("Recursion"),
    answer("Indexing"),
    allow_retry = TRUE
  )
)
```

---

## Making our own classes

While built-in classes are useful, we can also create our own. Classes are best
created using `structure()`. It's conventional to also write a **constructor
function** that validates input and builds the object for you.

```{r demo-custom-class, echo=TRUE}
# Constructor function for a "dna_sequence" class
dna_sequence <- function(seq) {
  # validate: only allow A, T, G, C characters
  valid <- all(strsplit(toupper(seq), "")[[1]] %in% c("A","T","G","C"))
  if (!valid) stop("Sequence must only contain A, T, G, C")

  structure(
    list(sequence = toupper(seq),
         length   = nchar(seq)),
    class = "dna_sequence"
  )
}

# Create an instance
my_gene <- dna_sequence("ATGCATGC")
my_gene
```

Now let's define a `print` method so it displays nicely:

```{r demo-print-method, echo=TRUE}
# Methods are named: generic.classname
print.dna_sequence <- function(x, ...) {
  cat("DNA Sequence\n")
  cat("  Length:", x$length, "bp\n")
  cat("  Seq:   ", x$sequence, "\n")
}

print(my_gene)
```

And a `summary` method that computes GC content:

```{r demo-summary-method, echo=TRUE}
summary.dna_sequence <- function(object, ...) {
  seq    <- strsplit(object$sequence, "")[[1]]
  gc     <- sum(seq %in% c("G","C")) / length(seq)
  cat("GC content:", round(gc * 100, 1), "%\n")
}

summary(my_gene)
```

### Your turn: explore the custom class

```{r ex-custom-class, exercise=TRUE, exercise.lines=10, exercise.setup="demo-summary-method"}
# Create a new dna_sequence object with the sequence "GCGCATTATGC"
# Then call print() and summary() on it
my_seq <- dna_sequence(___)
print(___)
summary(___)
```

```{r ex-custom-class-solution}
my_seq <- dna_sequence("GCGCATTATGC")
print(my_seq)
summary(my_seq)
```

### Your turn: build your own class

This is a stretch challenge — have a go, and revisit it later if it's too tricky right now.

```{r ex-build-class, exercise=TRUE, exercise.lines=30}
# Build a class called "species" that stores:
#   - name (character): the species name
#   - kingdom (character): "animal", "plant", "fungi", etc.
#   - endangered (logical): whether it is endangered

# 1. Write a constructor function `species()`
# 2. Write a print.species() method
# 3. Create an instance and print it

species <- function(name, kingdom, endangered) {
  structure(
    list(name = ___, kingdom = ___, endangered = ___),
    class = "species"
  )
}

print.species <- function(x, ...) {
  cat("Species:", ___, "\n")
  cat("Kingdom:", ___, "\n")
  cat("Endangered:", ___, "\n")
}

lion <- species("Panthera leo", "animal", TRUE)
print(lion)
```

```{r ex-build-class-solution}
species <- function(name, kingdom, endangered) {
  structure(
    list(name = name, kingdom = kingdom, endangered = endangered),
    class = "species"
  )
}

print.species <- function(x, ...) {
  cat("Species:", x$name, "\n")
  cat("Kingdom:", x$kingdom, "\n")
  cat("Endangered:", x$endangered, "\n")
}

lion <- species("Panthera leo", "animal", TRUE)
print(lion)
```

## Final quiz

```{r quiz-classes-final, echo=FALSE}
quiz(
  question("How do you name an S3 method for the `print` generic on a class called `myclass`?",
    answer("`print.myclass`", correct = TRUE),
    answer("`myclass.print`"),
    answer("`method_print_myclass`"),
    answer("`print@myclass`"),
    allow_retry = TRUE
  ),
  question("What function do you use to list all methods for a class?",
    answer("`methods(class = ...)`", correct = TRUE),
    answer("`list.methods(...)`"),
    answer("`class.methods(...)`"),
    answer("`show.methods(...)`"),
    allow_retry = TRUE
  ),
  question("What does `structure(obj, class = 'myclass')` do?",
    answer("Attaches a class attribute to obj", correct = TRUE),
    answer("Creates a new object from scratch"),
    answer("Checks if obj is the right type"),
    answer("Converts obj to a list"),
    allow_retry = TRUE
  )
)
```
