# Basic data types in R

All languages have data types, so R is no exception. We will look at all of the 
most common data types in this notebook.

## Primitive types

### Numbers

These are the most simple data types, and are used very commonly throughout all 
R code. First we have the numbers. All numbers in R without the `L` suffix are 
floating point numbers - aka decimals. They are of a 64 bit precision. These are
also known as `long` numbers or `double`s.

Typically this means we use:

* 53 bits for the coefficient (52 really)
* 11 for exponent
* 1 for sign (+/-)

So the largest number represented is:

```{r}
exponent <- 2^11 - 2^10 - 1
# roughly with some error
2 ^ exponent
```

And then you can have 16 digits precision after the decimal point:

```{r}
2^52
```

Anyways, here are some numbers!

```{r}
# all plain numbers in R are floats
x <- 1
# same thing
y <- 1.0
# same thing
z <- 1.

```

Integers are rarer in R. They are whole numbers. They can only be created by 
appending `L` to a number. They are of 32 bit precision. This discrepancy with
the above floats is actually quite annoying (see https://www.r-bloggers.com/2015/06/r-in-a-64-bit-world/).

This means in R, we can only assign numbers as high as 2^31 - 1 and as low as -2^31 - 1.
We take a bit for the +ve/-ve.

```{r}
# make the integer
my_integer <- 2L
# typeof is a function which shows the type of the object
typeof(my_integer)
```

### Boolean

Boolean types are true or false types. Internally they are treated as 1's or 0's.
They are binary types, so there are only two you can have. R calls these `logical`.

```{r}
# all caps true or false are the values
true <- TRUE
false <- FALSE

# you can short hand to T/F but I wouldn't
true_again <- T
not_true <- F

```

### Characters, or strings

R has strings, which are letters stuck together. These can be thought of as 
vectors of single letters, though R is rather annoying about strings. 

```{r}
a_string <- "hello, R"
typeof(a_string)
```

## Compound types

These types are a bit more complex than the primitive ones, and are composed of
them in different formats.

### Vectors

#### Atomic vectors

Vectors are extremely common in R, and you can't really be without them. They are
created (confusingly) using the `c()` function. This stands for concatenate.

```{r}
vec <- c(1, 2, 3, 4)
```

Importantly, vectors must contain all the same types within them. If you try to 
mix types up within a vector (which you can do), something called 'type coercion'
will happen.

```{r}
# number, number, character(!!), number, number
oops <- c(1, 2, "3", 4, 5)
oops
```

All the values in `oops` are now characters, as character takes precedence over
numbers. The full list of coercive precedence is here:

`character > complex > numeric > integer > logical > raw`

This means if I had a vector like this, the TRUE's will get changed into 1's.

```{r}
# uh oh mixing types again
bools <- c(TRUE, TRUE, 0)
bools
```

This behaviour is generally pretty undesirable, so try to avoid it if you can.
There are a bunch of functions for forcing different data types. For example:

```{r}
# force 1 into a character
char <- as.character(1)
char

# force the character "12" into an integer
int <- as.integer("12")
int
```

One of the powerful things about R is that many functions, such as `as.character()`
are applied across vectors by default. This is known as vectorisation. We'll probably
get to this later, as it's important.

```{r}
my_vector <- c(1, 2, 3, 4, 100)

as.character(my_vector)
```

#### Lists

Lists are different from vectors as they can contain elements of different types.
This makes them much more flexible. They can also contain elements of different
lengths - there are four 'slots' in this list below. The first three are of length
three, and the last has a single element.

```{r}
x <- list(1:3, c("a", "b", "c"), c(TRUE, TRUE, FALSE), 1)
x
```

You can even have nested lists. This makes them a bit like maps/hashmaps/associative arrays
in other languages.

```{r}
x <- list(list(1), list(list(2)))
x
```

Many more complicated data structures in R, like the fabled `data.frame` is actually
just a fancy list.

### Important special data types

There are two other data types I want to highlight here, matrices and data frames.
Data frames are perhaps the most widely used data types in R, so a good understanding
of these will help you a lot in your R quest.

#### Matrices

These are two dimensional vectors essentially; think of a table of numbers.
They can only be created with the `matrix` function.

```{r}
m <- matrix(1:6, ncol = 3, nrow = 2)
m
```

Notice the `1:6` is shorthand for a range, in this case for 1 to 6, i.e `c(1, 2, 3, 4, 5, 6)`.
Unless you are a maths/stats whiz, you probably won't come across these directly
too much. It is worth noting that matrices do have their own special functions.

```{r}
# you can specify the 'shape' of the matrix
# with a combination of the ncol and nrow arguments.
m <- matrix(1:6, ncol = 3, nrow = 2)
m2 <- matrix(1:6, ncol = 2, nrow = 3)

# matrix multiplication uses this jazzy notation.
m %*% m2
```

#### Data frames

You can create data frames like so, using the `data.frame` function.

```{r}
# make a data frame
d <- data.frame(x = 1:5, y = 4:8)
# check out its structure
str(d)

cat("The type of the data frame is actually:\n")
# the type of a data frame
typeof(d)
```

A *very useful function* is `str`, which stands for structure. It gives an overview
of the data structure, here we can see that there are two columns, both integers.

Notice that we assigned names to each of the columns, and we can refer to these
names later on. We'll look at this when we want to index into data frames.

Data frames can have mixed types in the columns, but they all must be the same
length.

## Over to you

1. What's the difference between a primitive and a compound type in R?
2. Vectors and lists are superficially similar. What's the difference?
3. Can you create a vector of integers?
4. Create a data frame with a numerical column, and a character column.



