---
title: "Basic Data Types in R"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 60)
```

## Primitive types

All languages have data types, so R is no exception. We will look at all of the
most common data types in this notebook.

### Numbers

These are the most simple data types, and are used very commonly throughout all
R code. All numbers in R without the `L` suffix are floating point numbers —
aka decimals. They are of 64 bit precision. These are also known as `double`s.

```{r demo-doubles, echo=TRUE}
# all plain numbers in R are floats (doubles)
x <- 1
y <- 1.0   # same thing
z <- 1.    # also the same thing

# check the type with typeof()
typeof(x)
```

Integers are rarer in R. They are whole numbers created by appending `L` to a
number. They are of 32 bit precision, so they max out at 2^31 − 1.

```{r demo-integer, echo=TRUE}
# make an integer
my_integer <- 2L
typeof(my_integer)
```

### Your turn: number types

```{r ex-numbers, exercise=TRUE, exercise.lines=6}
# 1. Create a floating point number called `body_mass` with value 3.2
# 2. Create an integer called `num_chromosomes` with value 46L
# 3. Print the typeof() for each

body_mass <- ___
num_chromosomes <- ___
typeof(body_mass)
typeof(num_chromosomes)
```

```{r ex-numbers-solution}
body_mass <- 3.2
num_chromosomes <- 46L
typeof(body_mass)
typeof(num_chromosomes)
```

```{r quiz-numbers, echo=FALSE}
quiz(
  question("What does the `L` suffix do when added to a number in R?",
    answer("Makes it an integer", correct = TRUE),
    answer("Makes it larger"),
    answer("Makes it a character"),
    answer("Makes it logical"),
    allow_retry = TRUE
  ),
  question("What is the default numeric type in R (no `L` suffix)?",
    answer("double", correct = TRUE),
    answer("integer"),
    answer("float"),
    answer("long"),
    allow_retry = TRUE
  )
)
```

### Boolean (logical)

Boolean types are `TRUE` or `FALSE`. Internally they are treated as 1s and 0s.
R calls these `logical`.

```{r demo-logical, echo=TRUE}
# all caps TRUE or FALSE
true_val  <- TRUE
false_val <- FALSE

# you can shorthand to T/F but I wouldn't recommend it
also_true <- T
```

```{r ex-logical, exercise=TRUE, exercise.lines=6}
# Create a logical variable called `is_mammal` set to TRUE
# Create a logical variable called `is_extinct` set to FALSE
# What is TRUE + TRUE in R? (Hint: booleans are secretly 0s and 1s)
is_mammal  <- ___
is_extinct <- ___
TRUE + TRUE
```

```{r ex-logical-solution}
is_mammal  <- TRUE
is_extinct <- FALSE
TRUE + TRUE  # gives 2 - booleans are 1s and 0s
```

### Characters (strings)

R has strings, which are letters stuck together.

```{r demo-character, echo=TRUE}
a_string <- "hello, R"
typeof(a_string)
```

```{r ex-character, exercise=TRUE, exercise.lines=5}
# Create a character variable called `species` containing "Homo sapiens"
# Print its type with typeof()
species <- ___
typeof(species)
```

```{r ex-character-solution}
species <- "Homo sapiens"
typeof(species)
```

---

## Compound types

These types are composed of the primitive ones in different formats.

### Atomic vectors

Vectors are extremely common in R and are created with the `c()` function
(which stands for **concatenate**).

```{r demo-vector, echo=TRUE}
vec <- c(1, 2, 3, 4)
vec
```

Importantly, vectors must contain all the same type. If you mix types, **type
coercion** happens — R converts everything to the most flexible type:

`character > complex > numeric > integer > logical > raw`

```{r demo-coercion, echo=TRUE}
# mixing a character into a numeric vector
oops <- c(1, 2, "3", 4, 5)
oops  # everything is now a character!

# mixing logical into numeric
bools <- c(TRUE, TRUE, 0)
bools  # TRUEs become 1s
```

You can force conversion with `as.*` functions:

```{r demo-conversion, echo=TRUE}
char   <- as.character(1)
char

int_val <- as.integer("12")
int_val
```

Vectorisation means functions apply to every element automatically:

```{r demo-vectorised, echo=TRUE}
my_vector <- c(1, 2, 3, 4, 100)
as.character(my_vector)
```

### Your turn: vectors

```{r ex-vector, exercise=TRUE, exercise.lines=8}
# 1. Create a numeric vector called `wing_lengths` with values 12.1, 15.3, 11.8, 14.0
# 2. Create a character vector called `species_names` with 3 bird species names of your choice
# 3. What happens when you combine them with c()? Try it.
wing_lengths  <- c(___)
species_names <- c(___)

# combine and observe coercion
c(wing_lengths, species_names)
```

```{r ex-vector-hint}
# c() takes comma-separated values
# wing_lengths <- c(12.1, 15.3, 11.8, 14.0)
```

```{r ex-vector-solution}
wing_lengths  <- c(12.1, 15.3, 11.8, 14.0)
species_names <- c("Robin", "Sparrow", "Finch")
c(wing_lengths, species_names)  # all become characters
```

```{r quiz-vectors, echo=FALSE}
quiz(
  question("What does `c()` stand for?",
    answer("Concatenate", correct = TRUE),
    answer("Create"),
    answer("Column"),
    answer("Calculate"),
    allow_retry = TRUE
  ),
  question("What happens when you put a character inside a numeric vector?",
    answer("All values become characters (type coercion)", correct = TRUE),
    answer("The character is ignored"),
    answer("R throws an error"),
    answer("The character becomes 0"),
    allow_retry = TRUE
  )
)
```

### Lists

Lists can contain elements of **different types and different lengths** — much more flexible than vectors.

```{r demo-list, echo=TRUE}
# a list with numeric, character, and logical elements
organism <- list(
  genome_size_mb = 3200,
  species        = "Homo sapiens",
  is_sequenced   = TRUE
)
organism
```

```{r ex-list, exercise=TRUE, exercise.lines=8}
# Create a list called `penguin` with:
# - species: "Adelie"
# - bill_length_mm: 39.1
# - island: "Torgersen"
# - is_adult: TRUE
penguin <- list(
  species        = ___,
  bill_length_mm = ___,
  island         = ___,
  is_adult       = ___
)
penguin
```

```{r ex-list-solution}
penguin <- list(
  species        = "Adelie",
  bill_length_mm = 39.1,
  island         = "Torgersen",
  is_adult       = TRUE
)
penguin
```

### Factors

Factors are categorical variables that can only take a specific set of values
(called **levels**). They're essential for statistics and plotting.

```{r demo-factor, echo=TRUE}
# months recorded, out of order
recorded_months <- c("Oct", "Jan", "Mar", "Sep")

# define the correct order
month_levels <- c("Jan","Feb","Mar","Apr","May","Jun",
                  "Jul","Aug","Sep","Oct","Nov","Dec")

# create factor and sort
f <- factor(recorded_months, levels = month_levels)
sort(f)
```

```{r ex-factor, exercise=TRUE, exercise.lines=8}
# Ecologists recorded three habitat types, out of order
# Create a factor with levels ordered from least to most disturbed:
# "pristine" < "degraded" < "urban"
habitats <- c("urban", "pristine", "degraded", "urban", "pristine")

habitat_levels <- c(___, ___, ___)

habitat_factor <- factor(habitats, levels = habitat_levels)
sort(habitat_factor)
```

```{r ex-factor-solution}
habitats       <- c("urban", "pristine", "degraded", "urban", "pristine")
habitat_levels <- c("pristine", "degraded", "urban")
habitat_factor <- factor(habitats, levels = habitat_levels)
sort(habitat_factor)
```

### Data frames

Data frames are the most important compound type in R. They are tables where
each column can have a different type, but all columns must be the same length.

```{r demo-dataframe, echo=TRUE}
d <- data.frame(x = 1:5, y = 4:8)
str(d)   # str() shows the structure - use this constantly!
typeof(d)  # a data frame is secretly a list underneath
```

```{r ex-dataframe, exercise=TRUE, exercise.lines=8}
# Create a data frame called `penguins_mini` with:
# - species: c("Adelie", "Chinstrap", "Gentoo")
# - mass_kg: c(3.5, 3.7, 5.1)
# - is_breeding: c(TRUE, FALSE, TRUE)
# Then call str() on it
penguins_mini <- data.frame(
  species    = c(___),
  mass_kg    = c(___),
  is_breeding = c(___)
)
str(penguins_mini)
```

```{r ex-dataframe-solution}
penguins_mini <- data.frame(
  species     = c("Adelie", "Chinstrap", "Gentoo"),
  mass_kg     = c(3.5, 3.7, 5.1),
  is_breeding = c(TRUE, FALSE, TRUE)
)
str(penguins_mini)
```

## Final quiz

```{r quiz-types-final, echo=FALSE}
quiz(
  question("What is the key difference between a vector and a list?",
    answer("Vectors hold one type; lists can hold mixed types", correct = TRUE),
    answer("Lists are faster than vectors"),
    answer("Vectors can be longer than lists"),
    answer("There is no difference"),
    allow_retry = TRUE
  ),
  question("Can a data frame have columns of different types?",
    answer("Yes, each column can be a different type", correct = TRUE),
    answer("No, all columns must be the same type"),
    answer("Only if you use as.data.frame()"),
    allow_retry = TRUE
  ),
  question("You have a vector `c(1, 2, 3)`. What type is it?",
    answer("double (numeric)", correct = TRUE),
    answer("integer"),
    answer("character"),
    answer("logical"),
    allow_retry = TRUE
  )
)
```
