---
title: "Variables in R"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 60)
```

## What is a variable?

Variables are the building blocks of code. They are a very simple statement of
assignment. They are really for setting apart a block of memory in your computer
so you can access the variable later. Use `<-` to assign a value to a variable.

```{r demo-assign, echo=TRUE}
# assigning a value 1 to x
x <- 1
# returning x
x
```

We can assign as much as we like.

```{r demo-overwrite, echo=TRUE}
# x is 1 and y is 2
x <- 1
y <- 2

# overwrite x to 5
x <- 5

# return x
x
```

Now we can start building a simple program.

```{r demo-program, echo=TRUE}
# assign x to 1 and y to 5
x <- 1
y <- 5

# multiply x and y together
z <- x * y

# the cat function prints text to the terminal
cat("The value of z is:", z)
```

### Your turn: assign a variable

Fill in the blank (`___`) to assign the number `7` to a variable called `num_species`.

```{r ex-assign, exercise=TRUE, exercise.lines=4}
# assign 7 to num_species
num_species <- ___
num_species
```

```{r ex-assign-hint-1}
# The assignment operator in R is <-
# num_species <- ???
```

```{r ex-assign-solution}
num_species <- 7
num_species
```

### Your turn: multiply two variables

Create two variables and multiply them together.

```{r ex-multiply, exercise=TRUE, exercise.lines=7}
# Create a variable called `gc_content` with value 0.52
# Create a variable called `genome_length` with value 3200000
# Multiply them and store the result in `gc_count`
gc_content <- ___
genome_length <- ___
gc_count <- ___
gc_count
```

```{r ex-multiply-hint-1}
# Use <- to assign each variable, then * to multiply
```

```{r ex-multiply-hint-2}
gc_content <- 0.52
genome_length <- 3200000
gc_count <- gc_content * genome_length
```

```{r ex-multiply-solution}
gc_content <- 0.52
genome_length <- 3200000
gc_count <- gc_content * genome_length
gc_count
```

## Quick check

```{r quiz-variables, echo=FALSE}
quiz(
  question("Which operator do you use to assign a variable in R?",
    answer("`<-`", correct = TRUE),
    answer("`=>`"),
    answer("`==`"),
    answer("`->`"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("What does this code print? `x <- 3; x <- 10; x`",
    answer("10", correct = TRUE, message = "Correct! The second assignment overwrites the first."),
    answer("3"),
    answer("13"),
    answer("An error"),
    allow_retry = TRUE
  ),
  question("What does the `cat()` function do?",
    answer("Prints text to the console", correct = TRUE),
    answer("Creates a vector"),
    answer("Assigns a variable"),
    answer("Counts elements"),
    allow_retry = TRUE
  )
)
```

## Explore further

4. Are there other ways to assign to variables in R? For help, type `` ?`<-` ``
   in the console and read the help page.

```{r ex-explore, exercise=TRUE, exercise.lines=5}
# Try using = for assignment (it works too, but is less idiomatic)
# Try using -> for right-to-left assignment
# Which style does R documentation use?

```

```{r ex-explore-solution}
# = also works for assignment
x = 5
x

# -> assigns in the opposite direction
10 -> y
y

# R convention strongly prefers <- for clarity
```
