---
title: "Basic Plots in R"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 60)

barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)
```

## Plotting in R

One of the great strengths of R are its plotting abilities. Here we run through
**base R** plotting. Base plots are excellent for quick exploratory work — the
code is very terse — and with some effort they can produce publication-quality
figures.

---

## Plotting vectors (1D data)

### Line plots

When the index of a vector is meaningful, plot the raw data as a line:

```{r demo-line, echo=TRUE}
# random sample: index represents sampling order
`Random values (0-1000)` <- sample(x = 1:1000, size = 200, replace = TRUE)
plot(`Random values (0-1000)`, type = "l")
```

### Histograms

Histograms show the frequency distribution of a numeric variable — very useful
for assessing whether data are approximately normal.

```{r demo-hist, echo=TRUE}
# 600 random normal values
hist_data <- rnorm(n = 600, mean = 2, sd = 1.2)
hist(hist_data)
```

### Your turn: histograms

```{r ex-hist, exercise=TRUE, exercise.lines=10}
# Generate 500 random values from an exponential distribution (rexp)
# with rate = 0.5, and plot a histogram
# Add a main title with main = "..." and an x-axis label with xlab = "..."

exp_data <- rexp(n = ___, rate = ___)
hist(exp_data, main = ___, xlab = ___)
```

```{r ex-hist-hint}
# rexp(n = 500, rate = 0.5) generates the data
# hist() plots it; main and xlab are optional but good practice
```

```{r ex-hist-solution}
exp_data <- rexp(n = 500, rate = 0.5)
hist(exp_data, main = "Exponential distribution (rate = 0.5)", xlab = "Value")
```

```{r quiz-plots-1, echo=FALSE}
quiz(
  question("What plot type is best for showing the distribution of a single continuous variable?",
    answer("Histogram", correct = TRUE),
    answer("Scatter plot"),
    answer("Bar plot"),
    answer("Line plot"),
    allow_retry = TRUE
  ),
  question("What does `type = 'l'` do in `plot()`?",
    answer("Draws lines instead of points", correct = TRUE),
    answer("Draws labels"),
    answer("Increases line width"),
    answer("Sets the axis limits"),
    allow_retry = TRUE
  )
)
```

---

## Bivariate data

### Scatter plots (numeric vs numeric)

```{r demo-scatter, echo=TRUE}
var1 <- runif(100)
var2 <- sqrt(var1) * runif(100)

plot(var1, var2,
     xlab = "Variable 1",
     ylab = "Variable 2",
     main = "Scatter plot example",
     pch  = 19,   # filled circles
     col  = "steelblue")
```

### Your turn: scatter plot with cats data

```{r ex-scatter, exercise=TRUE, exercise.lines=14}
cats_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/cats.csv"
cats <- read.csv(cats_url)

# Plot heart weight (Hwt) on the y-axis against body weight (Bwt) on the x-axis
# Add informative axis labels and a title
plot(
  x    = ___,
  y    = ___,
  xlab = "Body weight (kg)",
  ylab = ___,
  main = ___,
  pch  = 19,
  col  = "darkred"
)
```

```{r ex-scatter-solution}
cats_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/cats.csv"
cats <- read.csv(cats_url)
plot(
  x    = cats$Bwt,
  y    = cats$Hwt,
  xlab = "Body weight (kg)",
  ylab = "Heart weight (g)",
  main = "Cat body weight vs heart weight",
  pch  = 19,
  col  = "darkred"
)
```

### Bar plots (categorical data)

Bar plots are used when one variable is categorical. Here we tabulate letter
frequencies:

```{r demo-barplot, echo=TRUE}
freq_letters <- function(text) {
  text_as_vec <- unlist(strsplit(text, ""))
  tab         <- as.data.frame(table(text_as_vec))
  only_letters <- tab[tab$text_as_vec %in% letters, ]
  names(only_letters) <- c("Letter", "Frequency")
  rownames(only_letters) <- 1:nrow(only_letters)
  only_letters$Letter <- factor(only_letters$Letter, levels = letters)
  only_letters
}

freq_data <- freq_letters(as.character(list(freq_letters)))
barplot(Frequency ~ Letter, freq_data)
```

### Box plots (grouped data)

Box plots summarise a continuous variable across groups — excellent for
comparing distributions.

```{r demo-boxplot, echo=TRUE}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)

# par() controls graphics parameters
# las = 2 rotates x-axis labels; mar adjusts margins
old_par <- par(las = 2, mar = c(8, 5, 2, 1))
boxplot(yield ~ variety, data = barley,
        xlab = "", ylab = "Yield (bushels/acre)",
        main = "Barley yield by variety",
        col  = "wheat")
par(old_par)  # restore original parameters
```

### Your turn: box plot

```{r ex-boxplot, exercise=TRUE, exercise.lines=12}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)

# Plot yield by SITE (not variety)
# Rotate x-axis labels and add a meaningful y-axis label
old_par <- par(las = 2, mar = c(8, 5, 2, 1))
boxplot(___ ~ ___, data = barley,
        xlab = "", ylab = ___,
        main = "Barley yield by site",
        col  = "lightblue")
par(old_par)
```

```{r ex-boxplot-solution}
barley_url <- "https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv"
barley <- read.csv(barley_url)
old_par <- par(las = 2, mar = c(8, 5, 2, 1))
boxplot(yield ~ site, data = barley,
        xlab = "", ylab = "Yield (bushels/acre)",
        main = "Barley yield by site",
        col  = "lightblue")
par(old_par)
```

## Choosing the right plot

```{r quiz-plot-choice, echo=FALSE}
quiz(
  question("You want to compare the distribution of body weights across three species. Which plot type is most appropriate?",
    answer("Box plot", correct = TRUE),
    answer("Line plot"),
    answer("Scatter plot"),
    answer("Pie chart"),
    allow_retry = TRUE
  ),
  question("You want to show the relationship between two continuous variables (e.g. height and weight). Which plot type?",
    answer("Scatter plot", correct = TRUE),
    answer("Bar plot"),
    answer("Histogram"),
    answer("Box plot"),
    allow_retry = TRUE
  ),
  question("What does `par()` do?",
    answer("Sets or queries graphical parameters for the current plot device",
           correct = TRUE),
    answer("Creates a new plot"),
    answer("Adds a legend to a plot"),
    answer("Saves a plot to a file"),
    allow_retry = TRUE
  )
)
```

## Next steps

Base R plots are great for exploration. In the next module (**Advanced Plots**)
we will look at **ggplot2**, which produces more polished, publication-ready
figures with a consistent and elegant syntax.
