# Plotting in R

One of the great strengths of R are the plotting abilities. Here we will run through the `base` plotting. `base` plots are excellent for exploratory work (as the code to generate them is so terse), but as we will see perhaps later on, they can also produce publication quality plots.

## Plotting vectors (i.e. 1D data)

Vector plots can be useful to either plot the raw data along with an index (if the index is meaningful), or can be plotted as frequency distributions (e.g. to assess normality).

```{r}
# here the index is meaningful - it's the place at which we drew a particular sample
# note if variables are in back ticks, you can add spaces and all sorts.
# this is a numeric vector of random numbers from 1 to 1000
`Random values (0-1000)` <- sample(x = 1:1000, size = 200, replace = TRUE)

# just call plot - the type = "l" says draw lines. Dots are drawn by default
plot(`Random values (0-1000)`, type = "l")
```

Another usage is to somehow interpret a vector more than simply plotting the values as they occur. We may want to plot a histogram.

```{r}
# generate some random normal data with given mean and sd
# this is a numeric vector
hist_data <- rnorm(n = 600, mean = 2, sd = 1.2)

# call the hist function, it will bin and plot the data automatically for us
hist(hist_data)
```

## Plotting bivariate data

Bivariate simply means 'two variables'. It's the next step up from having a vector. It might be a `n x 2` column `matrix` or a data frame with two columns.

### Bivariate numeric data

It's a scatter plot generally.

```{r}
# 100 random uniformly distributed numbers
var1 <- runif(100)
# var2 as a function of var1
var2 <- sqrt(var1) * runif(100)

# make into data frame
vars <- data.frame(var1, var2)

# with allows us to specify the data once, and then use the column
# names as variables
plot(vars$var1, vars$var2)
```

### Bivariate numeric and categorical

This also happens a lot. We can take a simple example by calculating the frequencies of letters used in a particular text. A barplot is often a nice idea.

I introduce the `formula`

```{r}
# a little function to tabulate all letters from an input character
# string
freq_letters <- function(text) {
  # split the text
  text_as_vec <- unlist(strsplit(text, ""))
  # and tabulate
  tab <- as.data.frame(table(text_as_vec))
  
  # make sure we only have letters
  only_letters <- tab[tab$text_as_vec %in% letters, ]
  
  # rename columns and rownames
  names(only_letters) <- c("Letter", "Frequency")
  rownames(only_letters) <- 1:nrow(only_letters)
  # and make sure the letters are only letters as a factor
  only_letters$Letter <- factor(only_letters$Letter, levels = letters)
  only_letters
}

# a bit meta, but we can input the source code of `freq_letters`
# as the text we want to process
freq_letters_source <- as.character(list(freq_letters))
freq_data <- freq_letters(freq_letters_source)

# the actual plot call is very simple
barplot(Frequency ~ Letter, freq_data)
```

But if we have multiple observations per category, we might want to summarise the data somehow. Boxplots are a simple way of doing this. In this example I've used a few more of `plot`'s options. In particular, I've also introduced `par` which is a function to manipulate the graphics environment and takes a bunch of key/value (e.g. `x = 1`) pairs. We can assign this to a variable, then call `par` again on this variable to restore the default parameters.

I've edited the `las` (label style) to `2` to rotate the x axis labels around, and the `mar` (margins) of the plot.

```{r}
# load in the data
barley <- read.csv("https://raw.githubusercontent.com/ARU-life-sciences/bioinformatics/main/R/programming/2_working_with_dataframes/data/barley.csv")
# create the plot parameters
pbox <- par(las = 2, mar=c(8,8,1,1))
# plot the boxplot
boxplot(yield ~ variety, data = barley, xlab = "")
# restore default parameters
par(pbox)
```
