---
title: "Advanced Plots with ggplot2"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 60)

penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- read.csv(penguins_url)
penguins <- na.omit(penguins)  # drop rows with missing values for simplicity
```

## Why ggplot2?

Base R plots are great for quick exploration. **ggplot2** is great for
*everything else*. It produces beautiful, publication-ready figures with a
consistent, powerful grammar. It is one of the most downloaded R packages of all
time, and is the standard plotting tool in modern R.

Install it (run this once on your machine):

```{r install-ggplot, eval=FALSE}
install.packages("ggplot2")
```

```{r load-ggplot, eval=FALSE}
library(ggplot2)
```

---

## The grammar of graphics

ggplot2 is based on a **grammar of graphics**. Every plot is built from:

| Layer | What it is |
|-------|------------|
| **data** | The data frame containing your variables |
| **aes()** | Aesthetic mappings — which column maps to x, y, colour, size, etc. |
| **geom_*()** | Geometric objects — points, lines, bars, boxes, etc. |
| **labs()** | Labels — title, axis labels |
| **theme_*()** | The overall look of the plot |

The key insight: you build a plot by **adding layers** with `+`.

---

## Our dataset: Palmer Penguins

We will use the **Palmer Penguins** dataset — measurements of three penguin
species on islands near Palmer Station, Antarctica.

```{r demo-penguins, echo=TRUE}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- read.csv(penguins_url)
penguins <- na.omit(penguins)

str(penguins)
```

---

## Scatter plots: `geom_point()`

```{r demo-scatter, echo=TRUE}
ggplot(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point()
```

Add colour by species — just map `colour = species` in `aes()`:

```{r demo-scatter-colour, echo=TRUE}
ggplot(data = penguins,
       aes(x = bill_length_mm, y = bill_depth_mm, colour = species)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(
    title  = "Bill dimensions of three penguin species",
    x      = "Bill length (mm)",
    y      = "Bill depth (mm)",
    colour = "Species"
  ) +
  theme_bw()
```

### Your turn: scatter plot

```{r ex-scatter, exercise=TRUE, exercise.lines=12}
# Plot flipper_length_mm (x) vs body_mass_g (y)
# Colour points by island
# Add informative labels and theme_minimal()

ggplot(data = penguins,
       aes(x = ___, y = ___, colour = ___)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(
    title  = ___,
    x      = ___,
    y      = ___,
    colour = "Island"
  ) +
  ___()
```

```{r ex-scatter-hint}
# aes(x = flipper_length_mm, y = body_mass_g, colour = island)
# theme_minimal() is a clean theme
```

```{r ex-scatter-solution}
ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g, colour = island)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(
    title  = "Flipper length vs body mass by island",
    x      = "Flipper length (mm)",
    y      = "Body mass (g)",
    colour = "Island"
  ) +
  theme_minimal()
```

```{r quiz-scatter, echo=FALSE}
quiz(
  question("In ggplot2, what does `aes()` do?",
    answer("Maps data columns to visual properties (x, y, colour, size, etc.)",
           correct = TRUE),
    answer("Sets fixed colours for all points"),
    answer("Defines the axis limits"),
    answer("Adds axis labels"),
    allow_retry = TRUE
  ),
  question("How do you add layers in ggplot2?",
    answer("With the `+` operator", correct = TRUE),
    answer("With the `|>` pipe"),
    answer("By calling multiple plot functions"),
    answer("With commas inside ggplot()"),
    allow_retry = TRUE
  )
)
```

---

## Histograms: `geom_histogram()`

```{r demo-hist, echo=TRUE}
ggplot(data = penguins, aes(x = body_mass_g)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
  labs(title = "Distribution of penguin body mass",
       x = "Body mass (g)", y = "Count") +
  theme_bw()
```

Overlay histograms for each species using `fill = species` and `alpha` for
transparency:

```{r demo-hist-species, echo=TRUE}
ggplot(data = penguins, aes(x = body_mass_g, fill = species)) +
  geom_histogram(bins = 25, alpha = 0.6, position = "identity") +
  labs(title = "Body mass distribution by species",
       x = "Body mass (g)", y = "Count", fill = "Species") +
  theme_bw()
```

### Your turn: histogram

```{r ex-hist, exercise=TRUE, exercise.lines=10}
# Plot a histogram of flipper_length_mm, coloured by species
# Use alpha = 0.6 and position = "identity" for overlapping histograms

ggplot(data = penguins, aes(x = ___, fill = ___)) +
  geom_histogram(bins = 25, alpha = ___, position = "identity") +
  labs(title = "Flipper length distribution by species",
       x = "Flipper length (mm)", y = "Count") +
  theme_bw()
```

```{r ex-hist-solution}
ggplot(data = penguins, aes(x = flipper_length_mm, fill = species)) +
  geom_histogram(bins = 25, alpha = 0.6, position = "identity") +
  labs(title = "Flipper length distribution by species",
       x = "Flipper length (mm)", y = "Count") +
  theme_bw()
```

---

## Box plots: `geom_boxplot()`

```{r demo-boxplot, echo=TRUE}
ggplot(data = penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Body mass by species",
       x = "Species", y = "Body mass (g)") +
  theme_bw() +
  theme(legend.position = "none")  # legend is redundant here
```

Add the raw data points on top of the boxplot with `geom_jitter()`:

```{r demo-jitter, echo=TRUE}
ggplot(data = penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA) +
  geom_jitter(aes(colour = species), width = 0.2, size = 1, alpha = 0.6) +
  labs(title = "Body mass by species (with raw data)",
       x = "Species", y = "Body mass (g)") +
  theme_bw() +
  theme(legend.position = "none")
```

### Your turn: box plot

```{r ex-boxplot, exercise=TRUE, exercise.lines=12}
# Plot bill_length_mm by island, coloured by island
# Add jittered raw points on top

ggplot(data = penguins, aes(x = ___, y = ___, fill = ___)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA) +
  geom_jitter(aes(colour = ___), width = 0.2, size = 1.5, alpha = 0.6) +
  labs(title = "Bill length by island",
       x = "Island", y = "Bill length (mm)") +
  theme_bw() +
  theme(legend.position = "none")
```

```{r ex-boxplot-solution}
ggplot(data = penguins, aes(x = island, y = bill_length_mm, fill = island)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA) +
  geom_jitter(aes(colour = island), width = 0.2, size = 1.5, alpha = 0.6) +
  labs(title = "Bill length by island",
       x = "Island", y = "Bill length (mm)") +
  theme_bw() +
  theme(legend.position = "none")
```

---

## Bar plots: `geom_bar()` and `geom_col()`

`geom_bar()` counts rows automatically. `geom_col()` uses pre-computed heights.

```{r demo-bar, echo=TRUE}
# count penguins per species per island
ggplot(data = penguins, aes(x = island, fill = species)) +
  geom_bar(position = "dodge") +  # side-by-side bars
  labs(title = "Number of penguins per island by species",
       x = "Island", y = "Count", fill = "Species") +
  theme_bw()
```

### Your turn: bar plot

```{r ex-bar, exercise=TRUE, exercise.lines=10}
# Make a bar chart of penguin counts by sex, coloured by species
# Use position = "stack" for stacked bars

ggplot(data = penguins, aes(x = ___, fill = ___)) +
  geom_bar(position = ___) +
  labs(title = "Penguin counts by sex",
       x = "Sex", y = "Count", fill = "Species") +
  theme_bw()
```

```{r ex-bar-solution}
ggplot(data = penguins, aes(x = sex, fill = species)) +
  geom_bar(position = "stack") +
  labs(title = "Penguin counts by sex",
       x = "Sex", y = "Count", fill = "Species") +
  theme_bw()
```

---

## Faceting: small multiples

Faceting splits a plot into panels — one per level of a variable. This is one
of ggplot2's most powerful features.

```{r demo-facet, echo=TRUE}
ggplot(data = penguins,
       aes(x = bill_length_mm, y = bill_depth_mm, colour = sex)) +
  geom_point(size = 1.5, alpha = 0.7) +
  facet_wrap(~ species) +
  labs(title = "Bill dimensions by species and sex",
       x = "Bill length (mm)", y = "Bill depth (mm)",
       colour = "Sex") +
  theme_bw()
```

### Your turn: faceting

```{r ex-facet, exercise=TRUE, exercise.lines=12}
# Scatter plot of flipper_length_mm vs body_mass_g
# Colour by sex, and facet by ISLAND (not species this time)

ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g, colour = sex)) +
  geom_point(size = 1.5, alpha = 0.7) +
  facet_wrap(~ ___) +
  labs(title = "Flipper length vs body mass by island",
       x = "Flipper length (mm)", y = "Body mass (g)") +
  theme_bw()
```

```{r ex-facet-solution}
ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g, colour = sex)) +
  geom_point(size = 1.5, alpha = 0.7) +
  facet_wrap(~ island) +
  labs(title = "Flipper length vs body mass by island",
       x = "Flipper length (mm)", y = "Body mass (g)") +
  theme_bw()
```

---

## Adding trend lines: `geom_smooth()`

```{r demo-smooth, echo=TRUE}
ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g, colour = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +  # linear model with confidence band
  labs(title = "Flipper length vs body mass with linear trend",
       x = "Flipper length (mm)", y = "Body mass (g)",
       colour = "Species") +
  theme_bw()
```

---

## Saving plots

```{r demo-save, eval=FALSE}
# ggsave saves the last plot by default
# Change width, height, and dpi for different outputs

my_plot <- ggplot(data = penguins,
                  aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot() +
  theme_bw()

ggsave("penguins_boxplot.pdf", plot = my_plot, width = 8, height = 5)
ggsave("penguins_boxplot.png", plot = my_plot, width = 8, height = 5, dpi = 300)
```

## Final quiz

```{r quiz-ggplot-final, echo=FALSE}
quiz(
  question("Which ggplot2 geom would you use for a scatter plot?",
    answer("`geom_point()`", correct = TRUE),
    answer("`geom_scatter()`"),
    answer("`geom_dot()`"),
    answer("`geom_xy()`"),
    allow_retry = TRUE
  ),
  question("What does `facet_wrap(~ species)` do?",
    answer("Creates a separate panel for each level of `species`", correct = TRUE),
    answer("Colours points by species"),
    answer("Filters to only show species"),
    answer("Wraps the legend around the species names"),
    allow_retry = TRUE
  ),
  question("How do you add a linear trend line with a confidence band?",
    answer('`geom_smooth(method = "lm", se = TRUE)`', correct = TRUE),
    answer('`geom_line(method = "lm")`'),
    answer('`geom_trend(type = "linear")`'),
    answer('`stat_smooth(ci = TRUE)`'),
    allow_retry = TRUE
  ),
  question("In ggplot2, where do you define that point colour should vary by a variable?",
    answer("Inside `aes()`, e.g. `aes(colour = my_variable)`", correct = TRUE),
    answer("Inside `geom_point()`, e.g. `geom_point(colour = my_variable)`"),
    answer("In `labs(colour = my_variable)`"),
    answer("In `theme(colour = my_variable)`"),
    allow_retry = TRUE
  )
)
```
